### 2025-03-31 Literature-review/Preregistration/Preregistration-protocol-Taxometrics-NDD.Qmd
update

---
title: "Review protocol: A Critical Evaluation of Taxometric Analysis in the Study of Neurodevelopmental Disorders"
author: "Enrico Toffalini and Valentina Tobia"
date: 2025-03-31
format: 
    pdf
bibliography: Bibliography.bib
csl: assets/apa.csl
---

# 1. Objectives and rationale

In recent years, a transdiagnostic dimensional framework has gained increasing importance in the study of neurodevelopmental disorders, including learning disorders (e.g., dyslexia), attention-deficit/hyperactivity disorder (ADHD), and autism spectrum disorder (ASD) [@astle2022annual; @michelini2024neurodevelopmental]. This framework is consistent with the notions that neurodevelopmental disorders have a strong neurobiological and genetic basis (i.e., high heritability), and that the genes involved tend to be generalists, affecting multiple traits simultaneously [@plomin2005generalist]. More broadly, it reflects the understanding that complex behavioral traits are typically highly polygenic [@chabris2015fourth].

Debate over whether neurodevelopmental disorders should be understood as categorical entities has involved the use of data-driven methods such as clustering algorithms [e.g., @astle2022annual]. However, taxometric analysis appears to more directly address the issue. Unfortunately, it has been so far underused in this field of research. Taxometric analysis consists of a set of methods developed to determine whether the latent structure underlying psychological constructs, such as mental disorders, is categorical (i.e., "taxonic") or dimensional [@meehl1999clarifications; @meehl1996taxometric]. Although these methods have been extensively used in clinical psychology, and comprehensive reviews are now available [e.g., @haslam2020dimensions], their application in the context of neurodevelopmental disorders has been relatively limited and, possibly methodologically questionable. A recent debate involving @happe2020annual, @chown2021can, and @happe2021dimensional on the nature of autism discussed various lines of evidence regarding autism latent structure, but did not explicitly mention the potential of taxometric analysis. Other articles employing taxometric methods, as reviewed by @haslam2020dimensions, along with a recent empirical study by @frazier2023categorical, suggest that autism may be categorical. In contrast, ADHD appears to show a dimensional latent structure in taxometric investigations [@haslam2006latent; @marcus2011does]. Less is currently known about the latent structure of learning disorders. One article on dyslexia by @o2012taxometric suggested that already diagnosed children (by the way, using a collection of alternative criteria based on different reading indices) might be clustered into different subtypes, but it focused only on identifying subtypes and did not use taxometric analysis to examine whether dyslexia represents a taxon in relation to the general population.

One concern is that taxometric analysis is not only underused in the literature on neurodevelopmental disorders, but also that it is sometimes misapplied, leading to methodological issues. The most notable caveat, particularly evident in the above cited studies on autism, is the widespread use of artificial admixture. Artificial admixture refers to the practice of combining subgroups of participants selected using different criteria (e.g., a control group and a clinically diagnosed group) when conducting taxometric analysis. This practice is problematic because it can artificially produce a taxonic latent structure, even when the true underlying structure is dimensional [@grove1991validity; @ruscio2004conceptual].

The present review aims to critically evaluate the use of taxometric methods in published studies on neurodevelopmental disorders, with a particular focus on methodological issues, especially potential sources of bias such as artificial admixture and non-representative samples. The motivation is twofold. First, to assess whether the conclusions about latent structure drawn in these studies are supported by sound methodological practices. Second, to inform future research about the limitations and risks of applying taxometric analysis in this context. This is particularly urgent given the current state of knowledge on neurodevelopmental disorders and the increasing prominence of transdiagnostic dimensional approaches. Multiple lines of evidence (from behavioral genetics, psychometrics, and population-level studies) suggest that these conditions are best characterized as continuous and normally distributed. However, misapplications of taxometric methods may result in spurious claims of categorical boundaries, leading to confusion with significant implications for theory, clinical practice, and identity formation in children.

------------------------------------------------------------------------

# 2. Methods

## 2.1 Sample

### 2.1.1 Sources and Search strategy

A systematic search will be conducted in databases including PsycINFO, PubMed, Scopus, and Web of Science. The search will be performed on titles, abstracts, and keywords (excluding full texts), and may include the wildcard "*" to broaden the scope where supported (e.g., in Scopus):

(("taxometric" OR "taxometrics" OR "taxometric analysis" OR "MAMBAC" OR "MAXCOV" OR "MAXEIG" OR "L-Mode") AND ("autism" OR "ASD" OR "attention-deficit" OR "ADHD" OR "learning disorder" OR "learning disability" OR "dyslexia" OR "reading disability" OR "reading disorder" OR "dyscalculia" OR "mathematics disorder" OR "matehmatical learning difficulties"))

The precise query used for each search engine will be documented. Additional records will be identified through screening the reference lists of relevant systematic reviews [e.g., @haslam2020dimensions], citation tracking of key methodological articles, and supplementary searches via Google Scholar or AI-based tools such as OpenAI or Grok. The search will not be restricted by publication date but will be limited to published, peer-reviewed journal articles, as the primary aim of this review is to provide an overview of methodological limitations in the existing literature.

### 2.1.2 Eligibility Criteria

Inclusion criteria:

- The study applies at least one taxometric procedure (e.g., MAMBAC, MAXCOV, L-Mode, MAXEIG);

- The target population is individuals with neurodevelopmental disorders or traits;

- The study draws a conclusion about the latent structure (categorical vs. dimensional) or reports sufficient information for an independent reader to reach a conclusion based on those results (e.g., reports CCFIs).

Exclusion criteria:

- Theoretical, methodological, or simulation articles without empirical data;

- Articles not reporting any behavioral data (e.g., only neuroimaging or neurophysiological data);

- Reviews and/or meta-analyses;

- Studies using other latent structure methods only (e.g., LPA, mixture modeling) without any taxometric methods being employed.

## 2.2 Coding procedure

Two human coders will independently code each article and report the variables listed in Table 1. Any discrepancies will be resolved through discussion, either with the PI or, if the PI is one of the coders, with a third person. Since a single article may report multiple samples, the coded dataframe will follow a long-form data structure, with articles that may appear across multiple rows.

| **Variable Name** | **Description** | **Value Type** |
|:--------------------------|:-------------------------|:-----------|
| *ID_article* | A unique identifier of the article reporting the taxometric analysis. | Integer |
|  |  |  |
| *ID_sample* | A unique identifier of the sample within the article on which the taxometric analysis is performed. | Integer |
|  |  |  |
| *Authors* | Full list of the article authors' names. | Text |
|  |  |  |
| *Title* | Article title. | Text |
|  |  |  |
| *Year* | Year of publication of the article. | Integer |
|  |  |  |
| *Journal* | Name of the journal that published the article. | Text |
|  |  |  |
| *Target_disorder* | The neurodevelopmental condition addressed (e.g., Autistic Spectrum Disorder, ADHD, Dyslexia, Dyscalculia). | Category |
|  |  |  |
| *Population_type* | Type of population from which the sample was drawn (e.g., clinical, non-clinical, general unselected population, mixed/combined). | Category |
|  |  |  |
| *Sample_size* | Number of participants involved in the taxometric analysis | Integer |
|  |  |  |
| *Artificial_admixture* | Whether the analysis combined distinct subsamples separately drawn using alternative criteria (e.g., clinical + control; low score in a variable OR in another), possibly generating pseudotaxonic results. | Boolean |
|  |  |  |
| *Taxometric_methods_used* | Taxometric methods applied (e.g., MAMBAC, MAXCOV, L-Mode, MAXEIG). | List of categories |
|  |  |  |
| *Indicators_number* | Number of variables used as indicators in the taxometric analyses, if reported clearly. | Integer |
|  |  |  |
| *Indicators_skewness* | If reported, the skewness of the indicators. | List of numeric |
|  |  |  |
| *CCFI* | The value of the Comparison Curve Fit Index (CCFI), if reported. | Numeric |
|  |  |  |
| *Nuisance_covariance* | The within-group correlations, if reported separately by group. | List of numeric |
|  |  |  |
| *Indicators_validity* | The standardized mean difference (Cohen's d) between groups (taxon and complement). | Numeric |
|  |  |  |
| *Taxonic_conclusion* | Whether the authors explicitly concluded in favor of a taxonic structure. | Boolean |
|  |  |  |
| *Strength_of_conclusion* | If reported by the authors, the strength of the conclusion (e.g., strong, moderate, ambiguous). | Text |
|  |  |  |
| *Limits_reported* | Whether and what limitations are explicitly reported by the authors with regard to taxometric assumptions, sampling, or anything. | Text |

#### Use of AI tools

Generative AI tools (including GPT models from OpenAI) may be used to assist in locating relevant content within the reviewed articles. However, all coding decisions will be made by human researchers based on their own judgment. Fully automated coding process using available APIs might be conducted to explore the capability of AI in this context, but this will not be included in the review. If this is conducted, performance of AI will be reported and assessed against that of human coders. This automated coding will involve providing AI models different prompts to extract the same variables without human supervision. If conducted, this may serve to document on how accurately AI models can identify and extract methodological features in academic papers.

## 2.3 Data analysis

Data analysis will be primarily descriptive. Frequencies and proportions will be calculated for each coded variable. Patterns will be explored, with particular attention to the types of disorders studied; whether artificial admixture (i.e., combining clinical and control samples, or subsamples recruited using alternative criteria) was present; and the overall conclusions about latent structure. Specifically, we will examine whether methodological violations, particularly artificial admixture, high nuisance covariance, and low indicator validity, are systematically associated with conclusions in favor of taxonicity. Based on prior literature, studies that likely yield spurious evidence of taxonic structure will be flagged. A brief meta-summary will also be conducted, categorizing the strength of evidence for dimensional versus taxonic conclusions across studies for different types of neurodevelopmental disorders, using indices of methodological robustness, and especially the presence of artificial admixture, as moderators.

------------------------------------------------------------------------

# 3. Limitations

This review may not comprehensively address all possible methodological issues associated with taxometric methods. Many such issues exist and are more thoroughly discussed in the literature cited in the introductory section. These issues may be variously serious across different subfields of psychology. In the case of neurodevelopmental disorders, which are the focus of this review, sampling-related problems, particularly artificial admixture, appear especially critical. Another major concern is nuisance covariance, which indicates to excessive correlation among indicators within groups. In the literature on neurodevelopmental disorders, this might result from the uncontrolled use of multiple indicators that just reflect the same underlying construct (e.g., multiple measures of the same underlying reading ability). For this reason, nuisance covariance is also given specific attention here. Nonetheless, additional methodological issues may emerge and be investigated in future work, even if they are not included in the present preregistration.

This review does not address alternative methods for investigating categorical structure in neurodevelopmental disorders. Such methods include clustering approaches, such as latent profile analysis (LPA), latent class analysis (LCA), Gaussian mixture models, and other non-model-based clustering techniques, which are often used alongside taxometric methods in the relevant literature. An in-depth examination of these alternatives would require a much broader effort that may be undertaken in a future, more comprehensive review of the wider literature on clustering methods. Importantly, clustering approaches have their own assumptions, which, when violated, can frequently lead to false or misleading results in psychological research [@toffalini2022entia; @toffalini2024clusters]. For example, some methods assume local independence despite correlations across indicators (e.g., LPA, LCA, k-means algorithm when adapted for inferential purposes), while others assume normality of distributions (e.g., Gaussian mixture models, LPA). These issues, however, fall outside the scope of the current preregistered review.

------------------------------------------------------------------------

### Funding

This work is conducted as part of the project titled *"Tails or types? Testing the dimensional hypothesis in neurodevelopmental disorders"*, funded by the European Union - Next Generation EU, Mission 4, Component 1, CUP: C53D23004210006

------------------------------------------------------------------------

# References

### 2025-07-19 Paper/Paper.qmd
added partial draft of paper

---
title: "Tails or Types? A Critical Systematic Review of Taxometric Studies in Neurodevelopmental Disorders"
shorttitle: "Taxometric Review Neurodevelopmental Disorders"
authors:
  - name: Enrico Toffalini
    corresponding: true
    orcid: 0000-0002-1404-5133
    email: enrico.toffalini@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Margherita Calderan
    corresponding: false
    orcid: 0009-0005-5668-5162
    email: margherita.calderan@unipd.it
    affiliation: Department of General Psychology, University of Padua, Italy
  - name: Riccardo Pagan
    corresponding: false
    email: riccardo.pagan.3@studenti.unipd.it
    affiliation: Department of Social and Developmental Psychology, University of Padua, Italy
  - name: Valentina Tobia
    corresponding: false
    email: tobia.valentina@unisr.it
    affiliation: University Vita-Salute San Raffaele, Milan, Italy
author-note:
  disclosures:
    study-registration: 'The review was preregistered at https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0'
    financial-support: 'Financial support: Project 2022KBW99S "Tails or types? Testing the dimensional hypothesis in neurodevelopmental disorders", funded by Next Generation EU, Mission 4, Componente 1, CUP C53D23004210006'
    gratitude: "We are grateful to Athanasia Vasileiadou who first pointed us to some of the reviewed articles."
    data-sharing: null
    related-report: null
    conflict-of-interest: null
    authorship-agreements: null
abstract: "..."
keywords: [Neurodevelopmental disorders, Taxometric analysis, Autism, ADHD, Learning disorders]
bibliography: references.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-typst: default
  apaquarto-pdf:
    documentmode: man
---

In recent years, research on neurodevelopmental disorders (NDDs) has undergone widespread reconsiderations. Traditional diagnostic categories defined in the DSM-5 [@dsm5], have been increasingly called into question in favor of a dimensional and transdiagnostic view [e.g., @astle2022trandiagnostic; @michelini2024world; @happe2020looking; @happe2021both]. The DSM-5 itself has incorporated dimensional aspects, for example in defining a broad Autistic Spectrum Disorder (ASD) condition instead of a set of different autism-related diagnoses. Several arguments have contributed to this: the large heterogeneity within diagnostic groups, the poor alignment between data-driven clusters and DSM-based categories, the frequent comorbidity and phenotypic correlations across conditions, and the idea that individuals with NDDs might represent, at least in part, the tails of developmental continua, that are quantitatively rather than qualitatively different from the so-called neurotypical population [@michelini2024world; @happe2021both; @mammarella2021nocore].

Clarifying whether NDDs truly represent categories, rather than just tails of continuous distributions that span across the population, is important for different reasons. These involve quests about social and personal self-identity by involved individuals, discussions about diagnostic framing, comorbidity, and the choice on interventions in clinical settings, the structuring of classes, schools, and educational methods when dealing with neurodiversity, and applications to research design, planning of studies, and power analysis in research settings [@caviola2024editorial]. But most importantly, this is relevant for our understanding of the ontological status of reality, which should matter for scientific endeavor.

A dimensional approach fits well with long-standing principles in the study of individual differences in psychology, and has long characterized research on intelligence and personality [@caviola2024editorial]. It also seems appropriate for NDDs given accumulating evidence of polygenicity and multifactorial etiology [@plomin2005generalist; @kovas2006generalist; @pennington2012multiple; @astle2020core; @caviola2024editorial; @demontis2023adhdgenome]. In fact, most behavioral traits are believed to be influenced not by one or a few, but by many small, additive causes [@chabris2015fourth]. According to a famous principle known in statistics as the central limit theorem [@fischer2011central], when many individual causes (even discrete ones, e.g., gene variants) add up they converge to a joint distribution that tends to continuity and normality: this is, in fact, the "secret" of the Gaussian distribution. Nonetheless, a principled categorical view still resists in neurodevelopmental research, for example concerning ASD [e.g., @happe2021both; @chown2021bit; @frazier2023asd], subtypes within NDDs [e.g., @burgess2018taxodyslexia] or the possible emergence of new data-driven taxa [@astle2022trandiagnostic].

Adjudicating between whether a clinical diagnostic category is characterized by an underlying structure that is categorical or dimensional is precisely the focus of taxometric analysis, a set of quantitative methods that have been developed since the 1990s [e.g., @meehl1995taxometric]. Since then, taxometric analysis has been widely applied in psychopathology, generally providing evidence that most mental health conditions (e.g., depression, anxiety) are best conceptualized as dimensional rather than categorical [@haslam2012review; @haslam2020metaanalysis]. In brief, taxometric methods such as MAMBAC, MAXCOV, MAXEIG, and L-Mode, leverage patterns of covariation among observed indicators to test alternative competing latent models that are categorical or dimensional [@ruscio2007taxometric; @ruscio2004checklist; @cole2004taxometric]. Such methods have become increasingly refined through simulation-based techniques such as the Comparison Curve Fit Index (CCFI), which provide criteria for objective decision-making [@ruscio2018using].

Despite this, the use of taxometric analysis in neurodevelopmental research has been limited. Most existing studies focus on adult psychopathology, with few exceptions [@beauchaine2003taxometrics; @haslam2020metaanalysis]. @happe2021both and @chown2021bit in their debate on whether ASD is a category or a continuum not even mention the potential of taxometric analysis. In their recent influential article, @astle2022trandiagnostic briefly presents taxometry but gives it little emphasis, prevailing reviewing studies that employ cluster analysis. However, the limitations of clustering for taxometric purposes have been discussed for decades. Clustering algorithms tend to impose (rather than seek) categorical structure, often fail to recover known taxa in simulations, and generally provide weak tests of taxonicity [@beauchaine2003taxometrics]. Recent studies simulating realistic psychological data, characterized by a combination of (even modestly) skewed and correlated indicators, further suggest that common clustering algorithms easily detect pseudoclusters while having low power to test the existence of true taxa [@toffalini2022entia; @toffalini2024clusters]. Indeed, @astle2022trandiagnostic interpret results from clustering studies as evidence that data-driven structures fail to align with the expected DSM-5 categories, rather than as proper tests of taxonicity.

Taxometric analysis is not without limitations, however. Commonly cited threats to validity of taxometric conclusions include poor indicator validity, nuisance covariance, low base rate, and artificial admixture \[...\]. Poor indicator validity occurs when taxometric analysis is run on variables that present little mean differences between individuals with and without the target disorder (i.e., putatively, the taxon and the complement). Generally, it is recommended that Cohen's d ≫ 1 on each indicator. Nuisance covariance means that there is strong residual correlations after accounting for taxa (i.e., both within the taxon and the complement), thus reflecting the use of redundant indicators; residual r \> 0.40 is often regarded as problematic. Low base rate means that individuals with the taxonic condition are too few in the sample for taxometric analysis to correctly identify them; n \< 30 is often regarded as problematic. This is obviously a problem when the diagnostic category is rare (e.g., 1-2% of the population). All of these threats lead to the risk of failing to detect taxa where they exist. On the other side, artificial admixture consists of collecting participants using alternate recruitment channels. This is often (incorrectly) used to compensate for the low base rate, as many individuals with a diagnosis are recruited separately from the rest of the sample to increase the number of observations with the putative taxon. The resulting sample is an admixture of individuals all of whom might originally come from the same dimensional population, yet it gives the appearance that multiple subgroups are there.

In addition, many studies on autism have drawn categorical conclusions from data that likely suffer from artificial admixture—mixing separately recruited clinical and control samples—which is known to spuriously generate taxonic results even when the true latent structure is dimensional @grove1991; @ruscio2004clarifying; @toffalini2024airipa. While a handful of studies (e.g., @frazier2023categorical) suggest that autism may exhibit a categorical structure, these conclusions warrant skepticism given that they are typically based on admixed samples and rarely report key methodological parameters such as indicator validity or within-group covariance.

## Aims of the Present Review

# Methods

## Search Strategy and Study Selection

## Coding

very strong indicator skewness (for example due to using scales that only measure very rare and seriously problematic behaviors) might generate evidence for taxa where none exist.

# Results

# Discussion

The coexistence of nuisance covariance and/or poor indicator validity is practically unavoidable when true latent structure is dimensional. This is because, for an indicator to largely separate individuals with vs without a diagnosis, it must also be strongly related to the latent dimension, thus leading to strong nuisance covariance with any other indicator. But in fact, such empasse might directly reveal that the true structure of the putative disorder is dimensional.

While the DSM-5 may appear to promote a categorical model of disorders, it does not explicitly commit to a taxonic view. It is often researchers, not diagnostic manuals, who reify clinical categories as natural kinds, despite their originally pragmatic intent.

The fact that observed scores in psychology often deviate from normality [@micceri1989unicorn] may reflect psychometric characteristics (e.g., skewed indicators when observing times, accuracies, bounded sum scores) rather than the true shape of the underlying latent traits.

# References

::: {#refs}
:::

### 2025-07-20 Paper/Paper.qmd
paper update

@@
 
-Taxometric analysis is not without limitations, however.
+Taxometric analysis is not without limitations, however, some of which are shared with cluster analysis.
 Commonly cited threats to validity of taxometric conclusions include poor indicator validity, nuisance covariance, low base rate, and artificial admixture \[...\].
@@
 This is obviously a problem when the diagnostic category is rare (e.g., 1-2% of the population).
-All of these threats lead to the risk of failing to detect taxa where they exist.
-On the other side, artificial admixture consists of collecting participants using alternate recruitment channels.
-This is often (incorrectly) used to compensate for the low base rate, as many individuals with a diagnosis are recruited separately from the rest of the sample to increase the number of observations with the putative taxon.
-The resulting sample is an admixture of individuals all of whom might originally come from the same dimensional population, yet it gives the appearance that multiple subgroups are there.
+Finally, artificial admixture consists of mixing participants selected or recruited using alternate criteria.
+This is often (incorrectly) used to compensate for the low base rate, with many individuals with a diagnosis being recruited separately from the rest of the sample to increase the number of observations with the putative taxon.
+Poor indicator validity and low base rate risk masking detection of taxa, while strong nuisance covariance and artificial admixture risk to detecting pseudotaxa by mimicking categorical structures.
 
-In addition, many studies on autism have drawn categorical conclusions from data that likely suffer from artificial admixture—mixing separately recruited clinical and control samples—which is known to spuriously generate taxonic results even when the true latent structure is dimensional @grove1991; @ruscio2004clarifying; @toffalini2024airipa.
-While a handful of studies (e.g., @frazier2023categorical) suggest that autism may exhibit a categorical structure, these conclusions warrant skepticism given that they are typically based on admixed samples and rarely report key methodological parameters such as indicator validity or within-group covariance.
+## Goals of the Present Review
 
-## Aims of the Present Review
+The main goal of this systematic review is to critically examine the use of taxometric methods in the study of neurodevelopmental disorders (NDDs) within the published literature.
+We focused our review on the most commonly studied NDD categories: Autism Spectrum Disorder (ASD), Attention-Deficit/Hyperactivity Disorder (ADHD), and all forms of Specific Learning Disorders (SLDs), along with the closely related condition of Language Disorder/Impairment (the latter was not included in the preregistered protocol and thus constitutes a deviation).
+We excluded Intellectual Disability, as its etiological basis is often well identified and taxonic by definition in many cases (e.g., Down syndrome), or associated with identifiable medical risk factors.
+Residual non-syndromic cases, in turn, are thought to reflect the lower tail of general mental ability, and thus as inherently dimensional.
 
-# Methods
+A core aim of the review is to critically assess the methodological quality and credibility of findings reported in published taxometric studies.
+We evaluated potential sources of bias such as artificial admixture, and whether sufficiently complete methodological information was reported to judge other threats to validity, including low indicator validity, nuisance covariance, and the skewness of indicators.
+(The latter is relevant when scales assess rare or extreme behaviors, and can create the illusion of categorical structure).
 
-## Search Strategy and Study Selection
+Following this systematic review, we present a Monte Carlo simulation study to empirically demonstrate the extent to which artificial admixture can lead to false detection of categorical latent structures in dimensional data, both when using taxometric analysis and cluster analysis.
 
-## Coding
+# Systematic Review of Taxometric Analysis on NDDs
 
-very strong indicator skewness (for example due to using scales that only measure very rare and seriously problematic behaviors) might generate evidence for taxa where none exist.
+## Methods
 
-# Results
+The systematic review was conducted in accordance with the above listed goals, the preregistration protocol, and the PRISMA guidelines. The methods are detailed below.
+
+### Sources and Search strategy
+
+A systematic search was conducted on four databases: PsycINFO, PubMed, Scopus, Web of Science.
+No time limits were set, however virtually no records from before 1990 were expected, as taxometric methods were not yet developed or diffuse then.
+The following search query was performed on titles, abstracts, and/or keywords:
+
+((“taxometric\*” OR “MAMBAC” OR “MAXCOV” OR “MAXEIG” OR “L-Mode”) AND (“autis\*” OR “ASD” OR “attention-deficit\*” OR “ADHD” OR “learning dis\*” OR “reading dis\*” OR “dyslexia” OR “math\* dis\*" OR "math\* diff\*"))
+
+Where the wildcard "\*" was not supported (i.e., PubMed) it was replaced with full terms ("disorder", "disability" instead of "dis\*"; "mathematical", "mathematics", "math" instead of "math\*").
+
+Language disorder was subsequently added as a deviation from preregistration, replacing the second part of the above search query with ("language imp\*" OR "language dis\*").
+In addition, the full search query was launched on Google Scholar and the first 10 pages were screened; the references of and citations to @haslam2020metaanalysis were searched on Google Scholar for additional results; and a deep research via OpenAI's gpt-4o was launched providing it the preregistration protocol as context.
+All promising records thus emerging and not previously identified via databases, were listed under the *"Identification of studies via other methods"* section of the PRISMA flowchart.
+
+### Eligibility Criteria
+
+Inclusion criteria:
+
+-   The study applies at least one taxometric procedure (e.g., MAMBAC, MAXCOV, L-Mode, MAXEIG);
+
+-   The target population is individuals with NDDs or related traits;
+
+-   The study draws a conclusion about the latent structure (categorical vs.
+dimensional) or reports sufficient information for an independent reader to reach a conclusion based on those results (e.g., reports CCFIs).
+
+Exclusion criteria:
+
+-   Theoretical, methodological, or simulation articles without empirical data;
+
+-   Articles not reporting any behavioral data (e.g., only neuroimaging or neurophysiological data);
+
+-   Reviews and/or meta-analyses;
+
+-   Studies using other latent structure methods only (e.g., LPA, mixture modeling) without any taxometric methods being employed.
+
+### Screening and Coding
+
+Two independent human coders (MC and RP) screened and coded all articles.
+All discrepancies were resolved through discussion, and later revised by the PI (ET).
+No AI tools were used in the process, except in occasional cases by the human coders to assist them clarify the content of articles.
+
+Since a single article could report multiple (sub)samples and/or multiple taxometric analyses on the same sample, the full coded dataframe followed a long-form data structure, with some articles repeating over multiple rows.
+Coded variables are listed in the data dictionary table reported in the preregistration protocol, to which a few variables for notes and coders agreement were added.
+
+## Results
+
+# Simulation of Artificial Admixture
+
+We focus on artificial admixture because it is a very powerful threat to validity that could relatively easily be avoided by design when planning a study, and because as suggested by the result of the systematic review presented below, it is unfortunately frequent.
+There are different types of admixtures; @ruscio2004checklist discusses three cases.
+First, as mentioned above, the situation where clinical and nonclinical cases are separately recruited and merged.
+Second, the case of splitting a sample into subsamples (using alternative criteria) and then using those to test different taxometric questions.
+Third, trimming away observations of the putative complement (i.e., individuals without the disorder) to increase the base rate.
+All of these procedures artificially impose categorical structures to the data that do not exist in the population.
+It is worth mentioning that artificial admixture is also a threat for any type of cluster analysis, for the exact same reasons that it is a problem for taxometric analysis.
+
+We simulate a dimensional scenario that makes it visually compelling how artificial admixture of the first type mentioned above creates pseudotaxa.
+A single normally-distributed latent variable $X$ is generated, that defines the target condition.
+A series of observed indicators $x1, x2, etc.$ are simulated, incorporating the $X$ latent variable plus some random normally-distributed noise ($R^2 = 0.77$).
+Individuals above the 99^th^ percentile of $X$ receive a diagnosis via clinical assessment.
+A sample of $N = 1000$ individuals is collected from the general population.
+Since the base rate is low (on average, only about $n = 10$ cases are expected to naturally present a diagnosis), researchers select and recruit an additional $n = 1000$ cases with a clinical diagnosis from the general population, thus creating artificial admixture.
+All data simulation is available in the online materials at xxxxxxxxxxxxxxxx.
+
+An instance of the simulation, plotting two observed variables for illustrative purposes, is shown in @FIGURE_xx.
+Color is added for further clarity to distinguish individuals with and without a diagnosis, but two clusters still clearly emerge.
+Indicator validity is good (about $Cohen's$ $d = 2$), while nuisance covariance is partly problematic: for the clinical cases $r \approx 0$, but for the nonclinical cases $r = 0.61$.
+It is worth reminding that no taxa exist here, and the appearance of a categorical structure is mainly imposed by admixture, as shown below.
+
+A Monte Carlo simulation was then performed, with 1000 iterations.
+Taxometric analysis was conducted using the "RTaxometric" package of R.
+As recommended, CCFI was used for inference: CCFI varies between 0 (maximal evidence for dimensional structure) to 1 (maximal evidence for categorical structure), with 0.5 indicating inconclusive results.
+Performed procedures were, as per default: MAMBAC, MAXEIG, and L-Mode.
+A third $x3$ observed indicator (similar to those above) was entered to allow conducting all these taxometric procedures.
+Over 1000 simulated datasets, results always strongly favored a categorical structure: average combined CCFI = xxxx, 95% quantile interval \[xxxx, xxxx\] (average MAMBAC CCFI = xxxx; average MAXEIG CCFI= xxxx; average L-Mode CCFI = xxxx).
+The smallest combined CCFI over 1000 iterations was xxxx.
+To confirm that artificial admixture also leads to misleading results when performing clustering, Gaussian Mixture Models (GMM) was also run testing 1 vs 2 components, using the "mclust" package of R: over 1000 iterations, results always favored the 2-component relative to the 1-component model, with an average $\Delta BIC = xxxx$ in favor of the 2-component model.
+
+To confirm that the categorical structure was incorrectly favored specifically because of artificial admixture, we run another Monte Carlo simulation like the one above, but without admixture.
+In this case, a single sample of $N = 2000$ observations was drawn at each iteration.
+Therefore, nuisance covariance remains strong, but there is no admixture.
+Average combined CCFI = xxxx, 95% quantile interval \[xxxx, xxxx\]; average MAMBAC CCFI = xxxx; average MAXEIG CCFI= xxxx; average L-Mode CCFI = xxxx.
+Largest combined CCFI was xxxx.
+Thus, results remain frequently around inconclusive, probably due to nuisance covariance, but never strongly favor a categorical structure.
+GMM also consistently favored the 1-component model (over the 2-component model), with an average $\Delta BIC = xxxx$ in favor of the 1-component model.
+Therefore, admixture can be an extremely strong driver of false categorical findings, especially when many clinical cases are separately recruited and added to the sample.
 
@@
 
-The coexistence of nuisance covariance and/or poor indicator validity is practically unavoidable when true latent structure is dimensional.
-This is because, for an indicator to largely separate individuals with vs without a diagnosis, it must also be strongly related to the latent dimension, thus leading to strong nuisance covariance with any other indicator.
-But in fact, such empasse might directly reveal that the true structure of the putative disorder is dimensional.
+The present review suggests that taxometric analysis has been underused in neurodevelopmental research, despite its unique relevance for the recent debate on the categorical vs dimensional nature of disorders [@astle2022trandiagnostic; @michelini2024world; @happe2021both].
+Focusing on selected NNDs, we identified 14 relevant records.
+The underuse of taxometric analysis in this field was expected: @haslam2020metaanalysis identified 183 articles employing taxometric methods on psychopathology, of which only 15 concerned childhood (these included various NNDs but also other conditions such as sleep problems).
+What is most worrying, many of the articles that we reviewed present serious methodological limitations and/or incomplete reporting of relevant information.
+
+In the simulated example, artificial admixture was also accompanied by large nuisance covariance, although it was demonstrated that the latter posed a less severe problem to inference.
+In fact, the coexistence of low nuisance covariance and good indicator validity is practically impossible under a dimensional scenario.
+This is because an indicator can largely separate individuals with vs without a diagnosis if it is strongly related to the latent dimension, but this leads to nuisance covariance across valid indicators.
+Ultimately, however, such impasse might be interpreted as a direct signal that the true latent structure is dimensional.
 

### 2025-07-21 Literature-review/scripts/coded data processing.R
added data processing review

@@ -1,0 +1,15 @@
+
+# load libraries
+library(readxl)
+
+# import data
+df = data.frame(read_excel("../data/data_extraction.xlsx"))
+
+# coders agreement
+1 - mean(is.na(df$conclusion_agreement))
+mean(df$conclusion_agreement, na.rm=T)
+df$key = paste0(df$ID_article,df$ID_sample,df$subsample)
+x = aggregate(df$admixture_agreement, by=list(df$key), FUN=mean, na.rm=T)
+mean(x$x, na.rm=T)
+
+

### 2025-07-21 R/analysis-coded-review-data.R
R scripts

@@ -3,7 +3,7 @@
 library(readxl)
 
 # import data
-df = data.frame(read_excel("../data/data_extraction.xlsx"))
+df = data.frame(read_excel("Literature-review/data/data_extraction.xlsx"))
 
 # coders agreement
 1 - mean(is.na(df$conclusion_agreement))

### 2025-07-21 R/simulate-artificial-admixture.R

@@ -1,0 +1,130 @@
+
+################################################
+
+library(ggplot2)
+library(RTaxometrics)
+library(dplyr)
+library(mclust)
+
+################################################
+
+# visual demonstration
+
+set.seed(0)
+N = 1e7
+X = rnorm(N,0,1)
+x1 = scale(X*1.25 + rnorm(N,0,1))
+x2 = scale(X*1.25 + rnorm(N,0,1))
+x3 = scale(X*1.25 + rnorm(N,0,1))
+diagnosis = ifelse(X > quantile(X,0.99), 2, 1)
+population = data.frame(x1,x2,x3,diagnosis)
+
+df = population[1:1000,]
+df = rbind(df, population[diagnosis==2,][1000:1999,])
+df$Recruitment = c(rep("General unselected\npopulation",1000),rep("Added clinical\nsample",1000))
+df$Recruitment = relevel(as.factor(df$Recruitment),ref="General unselected\npopulation")
+
+ggplot(df, aes(x=x1,y=x2,color=Recruitment))+
+  scale_color_manual(values=c("#132b43","#46a1f7"))+
+  geom_point(size=3.5,alpha=.4)+
+  theme(text=element_text(size=25))+
+  xlab("x1 (z-score)")+ylab("x2 (z-score)")
+
+df |> group_by(Recruitment) |>
+  summarise(correlation=cor(x1,x2))
+
+################################################
+
+# monte carlo simulation with artificial admixture
+
+niter = 1000
+resultsAdmix = data.frame(CCFI.MAMBAC=rep(NA,niter),
+                     CCFI.MAXEIG=NA,
+                     CCFI.LMode=NA,
+                     CCFI.mean=NA,
+                     chosenG = NA,
+                     gmmBIC1 = NA,
+                     gmmBIC2 = NA,
+                     time=NA)
+
+N = 1e6
+for(i in 1:niter){
+  X = rnorm(N,0,1)
+  x1 = scale(X*1.25 + rnorm(N,0,1))
+  x2 = scale(X*1.25 + rnorm(N,0,1))
+  x3 = scale(X*1.25 + rnorm(N,0,1))
+  diagnosis = ifelse(X > quantile(X,0.99), 2, 1)
[…]

### 2025-07-21 Paper/Paper.qmd
update paper

@@
 
-The systematic review was conducted in accordance with the above listed goals, the preregistration protocol, and the PRISMA guidelines. The methods are detailed below.
+The systematic review was conducted according to the Preferred Reporting items for Systematic Reviews and Meta-Analysis 2020 guidelines.
+A review protocol was preregistered on OSF registries at the following link: https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0 The methods are detailed below.
 
@@
 
-A systematic search was conducted on four databases: PsycINFO, PubMed, Scopus, Web of Science.
-No time limits were set, however virtually no records from before 1990 were expected, as taxometric methods were not yet developed or diffuse then.
-The following search query was performed on titles, abstracts, and/or keywords:
+A systematic search was conducted for relevant peer-reviewed articles in April 2025 using four electronic databases: PsycINFO, PubMed, Scopus, Web of Science.
+No time limits were set, although no records were expected from before 1990, as taxometric methods were not yet developed or diffuse then.
+The following search query, reflecting the methods and the target populations of interest, was performed on titles, abstracts, and keywords:
 
@@
 
-Where the wildcard "\*" was not supported (i.e., PubMed) it was replaced with full terms ("disorder", "disability" instead of "dis\*"; "mathematical", "mathematics", "math" instead of "math\*").
+in databases where the wildcard "\*" was not supported (i.e., PubMed), it was replaced with full terms ("disorder", "disability" instead of "dis\*"; "mathematical", "mathematics", "math" instead of "math\*").
 
@@
 
-Inclusion criteria:
-
--   The study applies at least one taxometric procedure (e.g., MAMBAC, MAXCOV, L-Mode, MAXEIG);
-
--   The target population is individuals with NDDs or related traits;
-
--   The study draws a conclusion about the latent structure (categorical vs.
+Inclusion criteria were: 1) the study applies at least one taxometric procedure (e.g., MAMBAC, MAXCOV, L-Mode, MAXEIG); 2) the target population is individuals with NDDs or related traits; 3) the study draws a conclusion about the latent structure (categorical vs.
 dimensional) or reports sufficient information for an independent reader to reach a conclusion based on those results (e.g., reports CCFIs).
 
-Exclusion criteria:
-
--   Theoretical, methodological, or simulation articles without empirical data;
-
--   Articles not reporting any behavioral data (e.g., only neuroimaging or neurophysiological data);
-
--   Reviews and/or meta-analyses;
-
--   Studies using other latent structure methods only (e.g., LPA, mixture modeling) without any taxometric methods being employed.
+Exclusion criteria were: 1) the study only reports a theoretical, methodological, or simulation articles without empirical data; 2) the article does not report any behavioral data (e.g., only neuroimaging or neurophysiological data; this was later reconsidered, but no studies were excluded for this reason in any case); 3) the study only consists of a review and/or meta-analysis; 4) the study employs other latent structure methods only (e.g., LPA, mixture modeling), and not taxometric analysis.
+In addition, following the preregistration protocol, the search was limited to published, peer-reviewed journal articles, as a primary aim was to provide a critical review of the existing literature.
+However, one potentially interesting record excluded as non-journal article was separately discussed below in the Results section.
 
@@
 Two independent human coders (MC and RP) screened and coded all articles.
+No AI tools were used in the process, except in occasional cases by the human coders to assist them clarify the content of articles.
 All discrepancies were resolved through discussion, and later revised by the PI (ET).
-No AI tools were used in the process, except in occasional cases by the human coders to assist them clarify the content of articles.
 
 Since a single article could report multiple (sub)samples and/or multiple taxometric analyses on the same sample, the full coded dataframe followed a long-form data structure, with some articles repeating over multiple rows.
-Coded variables are listed in the data dictionary table reported in the preregistration protocol, to which a few variables for notes and coders agreement were added.
+Coded variables were listed in the data dictionary table defined in the preregistration protocol, to which a few variables for notes and coders agreement were added.
+
+Inter-coder agreement was fair to good in all phases.
+For the title+abstract screening phase, Cohen's k = 0.79 (34 out of 38 matching independent decisions); for the full-text eligibility screening, Cohen's k = 0.52 (15 out of 19).
+For final data extraction, 84% (71 out of 85) rows were extracted by both coders, with good agreement on the crucial information: for the "authors' taxonic conclusion" agreement was 94%, for artificial admixture was 95% at the sample/subsample level.
 
 ## Results
+
+A total of 110 records were identified through database searches.
+After removing 67 duplicates and excluding 5 records (dissertations or book sections, only one of which appeared to include primary taxometric analyses, but could not be retrieved), 38 unique articles were screened.
+Nineteen were excluded at the title/abstract level.
+Full-text assessment was conducted on the remaining 19 reports, with 7 excluded for not meeting inclusion criteria (4 did not use taxometric analysis; 3 did not target neurodevelopmental disorders).
+An additional 4 records were identified through citation tracking and online browsing, of which 2 were excluded (1 methodological article; 1 not employing taxometric methods).
+In total, 14 studies were included in the review.
+The PRISMA flowchart in @fig-prisma summarizes the selection process.
+
+@fig-prisma *about here*
+
+![PRISMA 2020 flow diagram illustrating the study screening and selection process](Figures/PRISMA_flow_diagram.png){#fig-prisma}
+
+## Additional Use of Cluster Analysis
+
+Among the 14 identified articles, six also employed methods of cluster analysis (in all but one case using latent class or latent profile analysis).
+In five of the six cases, cluster analysis was conducted in parallel with taxometric analysis [@james2016latent; @munson2008asd; @frazier2023asd; @frazier2007adhd; @deserno2023probing], while in one case it was performed as a secondary analysis [@frazier2010asd].
+In all cases, clustering results were interpreted as supporting taxonic conclusions, always converging with those from taxometric analyses, with the exception of @frazier2007adhd, where clustering results were partially inconclusive yet suggested the presence of subgroups, in contrast to the dimensional structure indicated by taxometric analysis.
+In all but two studies [@frazier2007adhd; @munson2008asd], we identified at least some risk of artificial admixture.
 
@@
 
+```{r}
+load("../R/simulateAdmixtResults.RData")
+```
+
 We simulate a dimensional scenario that makes it visually compelling how artificial admixture of the first type mentioned above creates pseudotaxa.
@@
 Since the base rate is low (on average, only about $n = 10$ cases are expected to naturally present a diagnosis), researchers select and recruit an additional $n = 1000$ cases with a clinical diagnosis from the general population, thus creating artificial admixture.
-All data simulation is available in the online materials at xxxxxxxxxxxxxxxx.
+All data simulation is available in the online materials at https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba
 
-An instance of the simulation, plotting two observed variables for illustrative purposes, is shown in @FIGURE_xx.
+An instance of the simulation, plotting two observed variables for illustrative purposes, is shown in @fig-admixture.
 Color is added for further clarity to distinguish individuals with and without a diagnosis, but two clusters still clearly emerge.
@@
 It is worth reminding that no taxa exist here, and the appearance of a categorical structure is mainly imposed by admixture, as shown below.
+
+@fig-admixture *about here*
+
+![A simulated example of artificial admixture, where a large number of individuals with a diagnosis are separately recruited and added to the sample (total N = 2000).
+The true underlying structure is strictly dimensional.](Figures/figureAdmixture.png){#fig-admixture}
 
@@
 A third $x3$ observed indicator (similar to those above) was entered to allow conducting all these taxometric procedures.
-Over 1000 simulated datasets, results always strongly favored a categorical structure: average combined CCFI = xxxx, 95% quantile interval \[xxxx, xxxx\] (average MAMBAC CCFI = xxxx; average MAXEIG CCFI= xxxx; average L-Mode CCFI = xxxx).
-The smallest combined CCFI over 1000 iterations was xxxx.
-To confirm that artificial admixture also leads to misleading results when performing clustering, Gaussian Mixture Models (GMM) was also run testing 1 vs 2 components, using the "mclust" package of R: over 1000 iterations, results always favored the 2-component relative to the 1-component model, with an average $\Delta BIC = xxxx$ in favor of the 2-component model.
+Over 1000 simulated datasets, results always strongly favored a categorical structure: median combined CCFI = `r formatC(median(resultsAdmix$CCFI.mean),2,format="f")`, 95% quantile interval \[`r formatC(quantile(resultsAdmix$CCFI.mean,.025),2,format="f")`, `r formatC(quantile(resultsAdmix$CCFI.mean,.975),2,format="f")`\]; median MAMBAC CCFI = `r formatC(median(resultsAdmix$CCFI.MAMBAC),2,format="f")`; average MAXEIG CCFI= `r formatC(median(resultsAdmix$CCFI.MAXEIG),2,format="f")`; average L-Mode CCFI = `r formatC(median(resultsAdmix$CCFI.LMode),2,format="f")`.
+The smallest combined CCFI over 1000 iterations was `r formatC(min(resultsAdmix$CCFI.mean),2,format="f")`.
+Therefore, admixture can be an extremely strong driver of false categorical findings, especially when many clinical cases are separately recruited and added to the sample.
+
+To confirm that artificial admixture leads to misleading results also when performing clustering, Gaussian Mixture Models (GMM) was also run testing 1 vs 2 components, using the "mclust" package of R.
+GMM is a type of finite mixture model, similar to latent profile analysis, but more flexible in that it also models indicator covariances.
+Over 1000 iterations, results always favored the 2-component relative to the 1-component model, with a median $\Delta BIC$ = `r formatC(median(resultsAdmix$gmmBIC1-resultsAdmix$gmmBIC2),2,format="f")` in favor of the 2-component model.
 
@@
 Therefore, nuisance covariance remains strong, but there is no admixture.
-Average combined CCFI = xxxx, 95% quantile interval \[xxxx, xxxx\]; average MAMBAC CCFI = xxxx; average MAXEIG CCFI= xxxx; average L-Mode CCFI = xxxx.
-Largest combined CCFI was xxxx.
-Thus, results remain frequently around inconclusive, probably due to nuisance covariance, but never strongly favor a categorical structure.
-GMM also consistently favored the 1-component model (over the 2-component model), with an average $\Delta BIC = xxxx$ in favor of the 1-component model.
-Therefore, admixture can be an extremely strong driver of false categorical findings, especially when many clinical cases are separately recruited and added to the sample.
+Median combined CCFI = `r formatC(median(resultsNoAdmix$CCFI.mean),2,format="f")`, 95% quantile interval \[`r formatC(quantile(resultsNoAdmix$CCFI.mean,.025),2,format="f")`, `r formatC(quantile(resultsNoAdmix$CCFI.mean,.975),2,format="f")`\]; median MAMBAC CCFI = `r formatC(median(resultsNoAdmix$CCFI.MAMBAC),2,format="f")`; average MAXEIG CCFI= `r formatC(median(resultsNoAdmix$CCFI.MAXEIG),2,format="f")`; average L-Mode CCFI = `r formatC(median(resultsNoAdmix$CCFI.LMode),2,format="f")`.
+Thus, results remain frequently around inconclusive, probably due to nuisance covariance, but do not risk strongly favoring a categorical structure.
+GMM also consistently favored the 1-component model (over the 2-component model), with an average $\Delta BIC$ = `r formatC(median(resultsNoAdmix$gmmBIC2-resultsNoAdmix$gmmBIC1),2,format="f")` in favor of the 1-component model.
 
@@
 
-The present review suggests that taxometric analysis has been underused in neurodevelopmental research, despite its unique relevance for the recent debate on the categorical vs dimensional nature of disorders [@astle2022trandiagnostic; @michelini2024world; @happe2021both].
-Focusing on selected NNDs, we identified 14 relevant records.
-The underuse of taxometric analysis in this field was expected: @haslam2020metaanalysis identified 183 articles employing taxometric methods on psychopathology, of which only 15 concerned childhood (these included various NNDs but also other conditions such as sleep problems).
-What is most worrying, many of the articles that we reviewed present serious methodological limitations and/or incomplete reporting of relevant information.
+The present review suggests that taxometric analysis has been underused in neurodevelopmental research, and that many articles using it present relevant methodological limitations or incomplete reporting of crucial information.
+This is unfortunate in light of the unique relevance of taxometric analysis for the current debate on the categorical vs dimensional nature of NDDs[@astle2022trandiagnostic; @michelini2024world; @happe2021both].
+The underuse of taxometric analysis in this field was expected: of the 183 articles employing taxometric methods on psychopathology identified by @haslam2020metaanalysis only 15 were placed in the "childhood" category (encompassing all types, even conditions such as sleep problems).
 

### 2025-07-21 R/analysis-coded-review-data.R

@@ -1,15 +1,22 @@
 
 # load libraries
 library(readxl)
+library(psych)
 
-# import data
+# import screening data
+scr = read.csv("Literature-review/data/processed/df_ok_screening.csv",sep=";")
+cohen.kappa(scr[,c("abs_coder1","abs_coder2")])
+sum(scr$abs_coder1==scr$abs_coder2); sum(!is.na(scr$abs_coder1))
+cohen.kappa(scr[,c("full_coder1","full_coder2")])
+sum(scr$full_coder1==scr$full_coder2,na.rm=T); sum(!is.na(scr$full_coder1))
+
+# import coded data
 df = data.frame(read_excel("Literature-review/data/data_extraction.xlsx"))
 
 # coders agreement
-1 - mean(is.na(df$conclusion_agreement))
+mean(!is.na(df$conclusion_agreement)); sum(!is.na(df$conclusion_agreement)); nrow(df)
 mean(df$conclusion_agreement, na.rm=T)
+df$conclusion_agreement
 df$key = paste0(df$ID_article,df$ID_sample,df$subsample)
 x = aggregate(df$admixture_agreement, by=list(df$key), FUN=mean, na.rm=T)
 mean(x$x, na.rm=T)
-
-

### 2025-07-21 R/simulate-artificial-admixture.R

@@ -5,6 +5,7 @@ library(ggplot2)
 library(RTaxometrics)
 library(dplyr)
 library(mclust)
+if(file.exists("R/simulateAdmixtResults.RData")) load("R/simulateAdmixtResults.RData")
 
 ################################################
 
@@ -27,6 +28,8 @@ df$Recruitment = relevel(as.factor(df$Recruitment),ref="General unselected\npopu
 ggplot(df, aes(x=x1,y=x2,color=Recruitment))+
   scale_color_manual(values=c("#132b43","#46a1f7"))+
   geom_point(size=3.5,alpha=.4)+
+  scale_x_continuous(breaks=seq(-10,10,1))+
+  scale_y_continuous(breaks=seq(-10,10,1))+
   theme(text=element_text(size=25))+
   xlab("x1 (z-score)")+ylab("x2 (z-score)")
 
@@ -78,7 +81,7 @@ for(i in 1:niter){
   
   print(resultsAdmix[i,])
 }
-save(resultsAdmix,file="R/artificialAdmixtResults.RData")
+save(resultsAdmix,file="R/simulateAdmixtResults.RData")
 
 ################################################
 
@@ -124,7 +127,7 @@ for(i in 1:niter){
   
   print(resultsNoAdmix[i,])
 }
-save(resultsNoAdmix,file="R/artificialNoAdmixtResults.RData")
+save(resultsAdmix,file="R/simulateAdmixtResults.RData")
 
 ################################################
 

### 2025-07-22 Paper/Paper.qmd
update paper

@@
     gratitude: "We are grateful to Athanasia Vasileiadou who first pointed us to some of the reviewed articles."
-    data-sharing: null
+    data-sharing: https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba
     related-report: null
@@
 
+```{r,message=F,warning=F}
+library(readxl)
+library(kableExtra)
+library(stringr)
+library(dplyr)
+```
+
 In recent years, research on neurodevelopmental disorders (NDDs) has undergone widespread reconsiderations.
-Traditional diagnostic categories defined in the DSM-5 [@dsm5], have been increasingly called into question in favor of a dimensional and transdiagnostic view [e.g., @astle2022trandiagnostic; @michelini2024world; @happe2020looking; @happe2021both].
+Traditional diagnostic categories defined in the DSM-5 [@dsm5], have been increasingly called into question in favor of a dimensional and transdiagnostic view [e.g., @astle2022trandiagnostic; @michelini2024world; @happe2020looking; @happe2021both; @sonuga2020kuhnian; @caviola2024editorial].
 The DSM-5 itself has incorporated dimensional aspects, for example in defining a broad Autistic Spectrum Disorder (ASD) condition instead of a set of different autism-related diagnoses.
@@
 
+```{r}
+df = data.frame(read_excel("../Literature-review/data/data_extraction.xlsx"))
+```
+
+An overview summary of the results (simplified to one row per article) is reported in @tbl-summary.
+A fully detailed reporting of coded data for all taxometric analyses is available in Supplemental online material.
+The total number of taxometric analyses was `r nrow(df)`, spanning from `r min(table(df$ID_article))` to `r max(table(df$ID_article))` per article.
+Targeted disorders were ADHD (6 articles), ASD (6 articles), Dyslexia (1 article), and Specific Language Impairment (SLI; 2 articles), with one article targeting both ADHD and ASD.
+Most articles (10 out of 14) focused on whether disorders were taxa, while the remaining 4 articles targeted whether subtypes were taxa.
+
+The most used taxometric method was MAMBAC, used in 13 out of 14 articles and in `r sum(grepl("MAMBAC",df$Taxometric_methods_used))` out of `r nrow(df)` total analyses, followed by MAXEIG (9 articles), L-Mode (7 articles), MAXCOV (2 articles), and MAXSLOPE (1 article).
+In most articles more than one taxometric method was used, with a median of `r median(str_count(df$Taxometric_methods_used,";")+1)` methods per analysis.
+In 11 out of 14 articles the interpretation was based on a quantitative index (Fit~d~ or, for more recent articles, CCFI), while in the remaining 3 cases the interpretation was based on visual inspection of taxometric curves.
+The number of indicators entered in the analyses spanned from `r min(df$Indicators_number,na.rm=T)` to `r max(df$Indicators_number,na.rm=T)`, with a median of `r median(df$Indicators_number,na.rm=T)` (excluding one article where the number was unclear).
+The median sample size across all analyses was `r median(df$Sample_size)`, or `r median((df |> group_by(ID_article) |> summarize(avg = mean(Sample_size)))$avg)` after averaging within article.
+
+Taxonic conclusions were mixed.
+All 5 articles targeting ADHD as a disorder concluded in favor of a dimensional structure, as did both articles targeting SLI.
+On the contrary, all 4 articles targeting ASD as a disorder leaned towards or clearly supported a categorical (taxonic) structure.
+The remaining 4 articles targeted subtypes of disorders, and all leaned towards taxonic conclusions, either strongly or predominantly [@stevens2018functional on ADHD subtypes; @munson2008asd on ASD subtypes; @obrien2012dyslexia on Dyslexia subtypes], or at least partly [@ingram2008defining, that supported ASD subtypes as taxa depending on what sets of indicators were used].
+In total, 8 out of 14 articles reached taxonic conclusions on at least some of their taxometric analyses (specifically, `r sum(df$Taxonic_conclusion)` out of `r nrow(df)` total analyses, or `r round(mean(df$Taxonic_conclusion)*100,1)`%, reached a taxonic conclusion), mostly concerning ASD (`r sum(grepl("ASD",df$Target_disorder)&df$Taxonic_conclusion==1)` out of `r sum(grepl("ASD",df$Target_disorder))` analyses, or `r round(100*sum(grepl("ASD",df$Target_disorder)&df$Taxonic_conclusion==1)/sum(grepl("ASD",df$Target_disorder)),1)`%).
+
+Artificial admixture was evident or strongly suspected, for at least part of the analyses, in 8 out of 14 articles (in total, involving `r sum(df$Artificial_admixture)` out of `r nrow(df)` analyses).
+Articles where artificial admixture was incurred tended to match those that reached taxonic conclusions, with 6 out of 8 cases.
+This matching was even higher at the level of individual analysis: taxonic conclusion was reached in `r sum(df$Taxonic_conclusion[df$Artificial_admixture==1])`out of `r sum(df$Artificial_admixture==1)` (`r round(100*mean(df$Taxonic_conclusion[df$Artificial_admixture==1]),1)`%) analyses where admixture was evident or suspected, in contrast to `r sum(df$Taxonic_conclusion[df$Artificial_admixture==0])` out of `r sum(df$Artificial_admixture==0)` (`r round(100*mean(df$Taxonic_conclusion[df$Artificial_admixture==0]),1)`%) analyses where it was probably or certainly excluded.
+
+Remaining methodological issues concerned missing information about descriptive statistics or other indices.
+In particular, nuisance covariances (i.e., correlations across indicators within diagnostic group or taxa) could be coded for only 2 out of 14 articles.
+Where reported, it was small to negligible in @james2016latent, while above 0.40 for many indicators in @frazier2023asd; both articles reached a taxonic conclusion.
+Skewness of distributions of indicators scores was reported by only 5 articles, and where reported it was often close to or beyond $\pm 1$ (at least one indicator in about one third of the analyses); this was not specifically associated to reaching a taxonic conclusion.
+On the contrary, indicators validity was largely reported, being coded for all but 3 articles, and it was almost always large or very large ($\gg1$).
+
+@tbl-summary *about here*
+
+```{r}
+#| label: tbl-summary
+#| tbl-cap: Overview summary of the reviewed articles."
+summary_tab = read_excel("Tables/Review-Summary.xlsx",sheet=2)
+kable(summary_tab, format = "latex", booktabs = TRUE) |>
+  kable_styling(font_size = 8, latex_options = c("striped", "scale_down")) |>
+  column_spec(1:ncol(summary_tab),width="5em") |>
+  landscape()
+```
+
 ## Additional Use of Cluster Analysis
 
-Among the 14 identified articles, six also employed methods of cluster analysis (in all but one case using latent class or latent profile analysis).
-In five of the six cases, cluster analysis was conducted in parallel with taxometric analysis [@james2016latent; @munson2008asd; @frazier2023asd; @frazier2007adhd; @deserno2023probing], while in one case it was performed as a secondary analysis [@frazier2010asd].
-In all cases, clustering results were interpreted as supporting taxonic conclusions, always converging with those from taxometric analyses, with the exception of @frazier2007adhd, where clustering results were partially inconclusive yet suggested the presence of subgroups, in contrast to the dimensional structure indicated by taxometric analysis.
-In all but two studies [@frazier2007adhd; @munson2008asd], we identified at least some risk of artificial admixture.
+Among the 14 identified articles, 6 also employed methods of cluster analysis (in all but one case using latent class or latent profile analysis).
+In 5 of the 6 cases, cluster analysis was conducted in parallel with taxometric analysis [@james2016latent; @munson2008asd; @frazier2023asd; @frazier2007adhd; @deserno2023probing], while in one case it was performed as a secondary analysis [@frazier2010asd].
+All 6 targeted ADHD and/or ASD, and in all cases clustering results were interpreted as supporting taxonic conclusions, always converging with those from taxometric analyses, with the exception of @frazier2007adhd, where clustering results were partially inconclusive yet suggested the presence of subgroups within ADHD, in contrast to the dimensional structure favored by taxometric analysis.
+In all but 2 studies [@frazier2007adhd; @munson2008asd], we identified at least some risk of artificial admixture.
 
@@
 
-The present review suggests that taxometric analysis has been underused in neurodevelopmental research, and that many articles using it present relevant methodological limitations or incomplete reporting of crucial information.
-This is unfortunate in light of the unique relevance of taxometric analysis for the current debate on the categorical vs dimensional nature of NDDs[@astle2022trandiagnostic; @michelini2024world; @happe2021both].
-The underuse of taxometric analysis in this field was expected: of the 183 articles employing taxometric methods on psychopathology identified by @haslam2020metaanalysis only 15 were placed in the "childhood" category (encompassing all types, even conditions such as sleep problems).
+The first finding of our critical review was that taxometric analysis has generally been underused in neurodevelopmental research, with only 14 articles being identified.
+This is unfortunate in light of the unique relevance of taxometric analysis for the current debate on the categorical vs dimensional nature of NDDs [@astle2022trandiagnostic; @michelini2024world; @happe2021both; @sonuga2020kuhnian].
+The underuse of taxometric analysis in this field was expected: in a meta-analysis conducted by @haslam2020metaanalysis, out 183 articles employing taxometric methods on psychopathology, only 15 were placed in the "childhood" category (this encompassed all types, even conditions such as sleep problems).
+Considering the studies covered in our review, only ASD, ADHD, and (to a minor extent) SLI were targeted as disorders in taxometric analysis.
+None of learning disorders was targeted as a taxon, with only Dyslexia being covered in one article, but only for investigating its subtypes.
+
+Despite some heterogeneity of methods, taxometric results were seemingly quite neat: dimensional conclusions were predominantly reached for ADHD and SLI, while categorical/taxonic conclusions were predominantly reached for ASD and wherever subtypes of any disorder were considered.
+This parallels a conclusion already advanced by @haslam2020metaanalysis, and seems to provide a strong answer to the debate by @chown2021bit and @happe2021both .
+However, a closer look at the methodological limitations of the reviewed studies warrants more caution in the conclusions.
+
+We identified four studies that targeted ASD as a disorder in taxometric analyses, and all of them reached taxonic conclusions.
+However, artificial admixture was identified in all of them.
+Artificial admixture consists of recruiting participants following alternate recruitment channels or criteria, and it is a major threat to conclusion validity, as shown in the above data simulation and in the literature \[...\].
+Admixture ranged from sampling participants through different methods, only some of which explicitly targeted disorders [@james2016latent], to admixing clinical and non-clinical samples [@frazier2023asd and @deserno2023probing; in the latter case inclusion criteria from a larger database are partly unclear, but the sample is explicitly structured by diagnosis] and composing a sample including affected and unaffected siblings [@frazier2010asd].
+According to our interpretation, the only unbiased taxometric evidence that ASD might be a taxon as a disorder might come from a secondary analysis by @frazier2010asd, where only unaffected siblings were examined, and a taxonic conclusion was still reached (due to the recruitment process, cases with a high level of the trait are oversampled here, but there is no explicit admixture).
+
+Concerning subtypes the situation was similar.
+All four reviewed articles partly or predominantly reached taxonic conclusions, but three of them probably or mostly incurred admixture.
+The issue was probably less serious here.
+In @obrien2012dyslexia, targeting Dyslexia subtypes, we identified only partly risk of admixture, as the authors used psychometric cutoffs on alternative tools and criteria for inclusion.
+This might not necessarily have affected the results, unless different criteria for inclusion correlate with different underlying deficits (phonological vs non-phonological impairments), however the authors themselves comment in their discussion that: "depending on the dyslexia criterion used, samples may vary in composition with regard to their subtypes" [@obrien2012dyslexia, p.33]
+
+Overall, at the article/sample level there were only 2 major mismatches: @marcus2012latent reached a dimensional conclusions despite incurring admixture by separately recruiting a large number of clinical cases; and @munson2008asd, who reached a taxonic conclusion about ASD subtypes despite no evidence of admixture (all children had received a diagnosis of ASD based on a seemingly uniform clinical assessment).
 

### 2025-07-22 R/1 analysis-review-data.R

@@ -1,0 +1,64 @@
+
+################################################################
+
+# load libraries
+
+library(readxl)
+library(psych)
+library(dplyr)
+library(stringr)
+
+################################################################
+
+# import data
+
+scr = read.csv("Literature-review/data/processed/df_ok_screening.csv",sep=";")
+df = data.frame(read_excel("Literature-review/data/data_extraction.xlsx"))
+
+################################################################
+
+# coders agreement
+
+# agreement on screening data
+cohen.kappa(scr[,c("abs_coder1","abs_coder2")])
+sum(scr$abs_coder1==scr$abs_coder2); sum(!is.na(scr$abs_coder1))
+cohen.kappa(scr[,c("full_coder1","full_coder2")])
+sum(scr$full_coder1==scr$full_coder2,na.rm=T); sum(!is.na(scr$full_coder1))
+
+# agreement on coded data
+mean(!is.na(df$conclusion_agreement)); sum(!is.na(df$conclusion_agreement)); nrow(df)
+mean(df$conclusion_agreement, na.rm=T)
+df$conclusion_agreement
+df$key = paste0(df$ID_article,df$ID_sample,df$subsample)
+x = aggregate(df$admixture_agreement, by=list(df$key), FUN=mean, na.rm=T)
+mean(x$x, na.rm=T)
+
+################################################################
+
+# statistics on coded data
+
+# total number of taxometric analyses, min and max per article
+nrow(df)
+tb=table(df$ID_article); min(tb); max(tb)
+
+# median sample size across all analyses
+median(df$Sample_size)
+
+# median sample size by article (averaged)
+median((df |> group_by(ID_article) |> summarize(avg = mean(Sample_size)))$avg)
+
+# median number of indicators per analysis
+median(df$Indicators_number,na.rm=T)
+
+# methods used
+grepl("MAMBAC",df$Taxometric_methods_used)
+sc = str_count(df$Taxometric_methods_used,";")+1; min(sc); median(sc); max(sc); mean(sc>1)
+
+# artificial admixture matching taxonic conclusions
+sum(df$Artificial_admixture)
+sum(df$Taxonic_conclusion)
+sum(df$Taxonic_conclusion[df$Artificial_admixture])
+
+################################################################
+
+

### 2025-07-22 R/analysis-coded-review-data.R

@@ -1,22 +1,0 @@
-
-# load libraries
-library(readxl)
-library(psych)
-
-# import screening data
-scr = read.csv("Literature-review/data/processed/df_ok_screening.csv",sep=";")
-cohen.kappa(scr[,c("abs_coder1","abs_coder2")])
-sum(scr$abs_coder1==scr$abs_coder2); sum(!is.na(scr$abs_coder1))
-cohen.kappa(scr[,c("full_coder1","full_coder2")])
-sum(scr$full_coder1==scr$full_coder2,na.rm=T); sum(!is.na(scr$full_coder1))
-
-# import coded data
-df = data.frame(read_excel("Literature-review/data/data_extraction.xlsx"))
-
-# coders agreement
-mean(!is.na(df$conclusion_agreement)); sum(!is.na(df$conclusion_agreement)); nrow(df)
-mean(df$conclusion_agreement, na.rm=T)
-df$conclusion_agreement
-df$key = paste0(df$ID_article,df$ID_sample,df$subsample)
-x = aggregate(df$admixture_agreement, by=list(df$key), FUN=mean, na.rm=T)
-mean(x$x, na.rm=T)

### 2025-07-23 Paper/Paper.qmd
update paper

--- 2025-07-20 Paper/Paper.qmd
@@
     authorship-agreements: null
-abstract: "..."
+abstract: "Taxometric methods are designed to test whether latent structures underlying psychological constructs are categorical (taxonic) or dimensional.
+Despite their relevance to the nature of neurodevelopmental disorders (NDDs), their application has been limited and often methodologically flawed.
+We conducted a preregistered systematic review to critically evaluate this literature.
+The review identified all published taxometric studies on autism, ADHD, learning, and language disorders, with a focus on methodological biases such as artificial admixture.
+While studies on ADHD and SLI supported dimensional structures, most studies on autism and NDD subtypes reported taxonic findings, often in the presence of methodological threats.
+A Monte Carlo simulation illustrates how artificial admixture can spuriously generate taxonic results in otherwise dimensional data.
+We conclude that evidence for categorical models in NDDs is frequently undermined by methodological artifacts, reinforcing the need for caution, but also advocating for more, high-quality taxometric studies in the field of NDDs."
 keywords: [Neurodevelopmental disorders, Taxometric analysis, Autism, ADHD, Learning disorders]
@@
 
-In recent years, research on neurodevelopmental disorders (NDDs) has undergone widespread reconsiderations.
-Traditional diagnostic categories defined in the DSM-5 [@dsm5], have been increasingly called into question in favor of a dimensional and transdiagnostic view [e.g., @astle2022trandiagnostic; @michelini2024world; @happe2020looking; @happe2021both].
-The DSM-5 itself has incorporated dimensional aspects, for example in defining a broad Autistic Spectrum Disorder (ASD) condition instead of a set of different autism-related diagnoses.
-Several arguments have contributed to this: the large heterogeneity within diagnostic groups, the poor alignment between data-driven clusters and DSM-based categories, the frequent comorbidity and phenotypic correlations across conditions, and the idea that individuals with NDDs might represent, at least in part, the tails of developmental continua, that are quantitatively rather than qualitatively different from the so-called neurotypical population [@michelini2024world; @happe2021both; @mammarella2021nocore].
+```{r,message=F,warning=F}
+library(readxl)
+library(kableExtra)
+library(stringr)
+library(dplyr)
+```
 
-Clarifying whether NDDs truly represent categories, rather than just tails of continuous distributions that span across the population, is important for different reasons.
-These involve quests about social and personal self-identity by involved individuals, discussions about diagnostic framing, comorbidity, and the choice on interventions in clinical settings, the structuring of classes, schools, and educational methods when dealing with neurodiversity, and applications to research design, planning of studies, and power analysis in research settings [@caviola2024editorial].
-But most importantly, this is relevant for our understanding of the ontological status of reality, which should matter for scientific endeavor.
+In recent years, research on neurodevelopmental disorders (NDDs) has undergone significant reconsideration.
+Traditional diagnostic categories defined in the DSM-5 [@dsm5] have increasingly been questioned in favor of a dimensional and transdiagnostic perspective [e.g., @astle2022trandiagnostic; @michelini2024world; @happe2020looking; @happe2021both; @sonuga2020kuhnian; @caviola2024editorial].
+Notably, the DSM-5 itself has incorporated some dimensional features (for example by defining a broad Autism Spectrum Disorder [ASD] in place of distinct diagnoses).
+Several factors have motivated this shift: the marked heterogeneity within diagnostic groups, the weak correspondence between data-driven clusters and DSM-based categories, the frequent comorbidity and phenotypic correlations across conditions, and the view that individuals with NDDs may represent, at least in part, the tails of developmental continua, quantitatively rather than qualitatively different from the so-called neurotypical population [@michelini2024world; @happe2021both; @mammarella2021nocore].
 
-A dimensional approach fits well with long-standing principles in the study of individual differences in psychology, and has long characterized research on intelligence and personality [@caviola2024editorial].
-It also seems appropriate for NDDs given accumulating evidence of polygenicity and multifactorial etiology [@plomin2005generalist; @kovas2006generalist; @pennington2012multiple; @astle2020core; @caviola2024editorial; @demontis2023adhdgenome].
-In fact, most behavioral traits are believed to be influenced not by one or a few, but by many small, additive causes [@chabris2015fourth].
-According to a famous principle known in statistics as the central limit theorem [@fischer2011central], when many individual causes (even discrete ones, e.g., gene variants) add up they converge to a joint distribution that tends to continuity and normality: this is, in fact, the "secret" of the Gaussian distribution.
-Nonetheless, a principled categorical view still resists in neurodevelopmental research, for example concerning ASD [e.g., @happe2021both; @chown2021bit; @frazier2023asd], subtypes within NDDs [e.g., @burgess2018taxodyslexia] or the possible emergence of new data-driven taxa [@astle2022trandiagnostic].
+Determining whether NDDs truly represent categories, rather than just tails of continuous distributions spanning the general population, is important for several reasons, including questions of social and personal identity for affected individuals, implications for diagnostic framing, comorbidity, and intervention choices in clinical contexts, and consequences for the organization of classrooms, educational practices, and even the design and power analysis of research studies [@caviola2024editorial].
+But more fundamentally, this question is relevant for our understanding of the ontological status of psychological phenomena.
 
-Adjudicating between whether a clinical diagnostic category is characterized by an underlying structure that is categorical or dimensional is precisely the focus of taxometric analysis, a set of quantitative methods that have been developed since the 1990s [e.g., @meehl1995taxometric].
-Since then, taxometric analysis has been widely applied in psychopathology, generally providing evidence that most mental health conditions (e.g., depression, anxiety) are best conceptualized as dimensional rather than categorical [@haslam2012review; @haslam2020metaanalysis].
-In brief, taxometric methods such as MAMBAC, MAXCOV, MAXEIG, and L-Mode, leverage patterns of covariation among observed indicators to test alternative competing latent models that are categorical or dimensional [@ruscio2007taxometric; @ruscio2004checklist; @cole2004taxometric].
-Such methods have become increasingly refined through simulation-based techniques such as the Comparison Curve Fit Index (CCFI), which provide criteria for objective decision-making [@ruscio2018using].
+A dimensional approach aligns well with long-standing principles in the study of individual differences in psychology, and has long characterized research on intelligence and personality [@caviola2024editorial].
+It also seems well-suited for NDDs, given a growing body of evidence supporting polygenicity and multifactorial etiology [@plomin2005generalist; @kovas2006generalist; @pennington2012multiple; @astle2020core; @caviola2024editorial; @demontis2023adhdgenome].
+Most behavioral traits are understood to result not from one or a few causes, but from the additive effects of many small influences [@chabris2015fourth].
+The central limit theorem [@fischer2011central] posits that, when many individual causes (such as discrete genetic variants) contribute additively, the resulting distribution tends to continuity and normality (this is, in fact, the "secret" of the Gaussian distribution).
+Nonetheless, a principled categorical view still resists in neurodevelopmental research, for example concerning ASD [e.g., @happe2021both; @chown2021bit; @frazier2023asd], subtypes within NDDs [e.g., @burgess2018taxodyslexia] or the proposed emergence of data-driven taxa [@astle2022trandiagnostic].
 
-Despite this, the use of taxometric analysis in neurodevelopmental research has been limited.
-Most existing studies focus on adult psychopathology, with few exceptions [@beauchaine2003taxometrics; @haslam2020metaanalysis].
-@happe2021both and @chown2021bit in their debate on whether ASD is a category or a continuum not even mention the potential of taxometric analysis.
-In their recent influential article, @astle2022trandiagnostic briefly presents taxometry but gives it little emphasis, prevailing reviewing studies that employ cluster analysis.
-However, the limitations of clustering for taxometric purposes have been discussed for decades.
-Clustering algorithms tend to impose (rather than seek) categorical structure, often fail to recover known taxa in simulations, and generally provide weak tests of taxonicity [@beauchaine2003taxometrics].
-Recent studies simulating realistic psychological data, characterized by a combination of (even modestly) skewed and correlated indicators, further suggest that common clustering algorithms easily detect pseudoclusters while having low power to test the existence of true taxa [@toffalini2022entia; @toffalini2024clusters].
-Indeed, @astle2022trandiagnostic interpret results from clustering studies as evidence that data-driven structures fail to align with the expected DSM-5 categories, rather than as proper tests of taxonicity.
+Taxometric analysis is a set of quantitative methods originally introduced by Paul Meehl and colleagues in the 1970s and refined over decades [e.g., @meehl1995taxometric], explicitly aimed at establishing whether a clinical disorder is characterized by an underlying categorical or dimensional structure.
+It has been widely applied in psychopathology, and typically suggested that most mental health conditions (e.g., depression, anxiety) are better conceptualized as dimensional [@haslam2012review; @haslam2020metaanalysis].
+In brief, taxometric procedures, such as MAMBAC, MAXCOV, MAXEIG, and L-Mode examine covariation patterns among observed indicators to compare competing latent models: one categorical, the other dimensional [@ruscio2007taxometric; @ruscio2004checklist; @cole2004taxometric].
+These methods now feature simulation-based tools like the Comparison Curve Fit Index (CCFI), which provide a quantitative measures to facilitate inference [@ruscio2018using].
 
-Taxometric analysis is not without limitations, however, some of which are shared with cluster analysis.
-Commonly cited threats to validity of taxometric conclusions include poor indicator validity, nuisance covariance, low base rate, and artificial admixture \[...\].
-Poor indicator validity occurs when taxometric analysis is run on variables that present little mean differences between individuals with and without the target disorder (i.e., putatively, the taxon and the complement).
-Generally, it is recommended that Cohen's d ≫ 1 on each indicator.
-Nuisance covariance means that there is strong residual correlations after accounting for taxa (i.e., both within the taxon and the complement), thus reflecting the use of redundant indicators; residual r \> 0.40 is often regarded as problematic.
-Low base rate means that individuals with the taxonic condition are too few in the sample for taxometric analysis to correctly identify them; n \< 30 is often regarded as problematic.
-This is obviously a problem when the diagnostic category is rare (e.g., 1-2% of the population).
-Finally, artificial admixture consists of mixing participants selected or recruited using alternate criteria.
-This is often (incorrectly) used to compensate for the low base rate, with many individuals with a diagnosis being recruited separately from the rest of the sample to increase the number of observations with the putative taxon.
-Poor indicator validity and low base rate risk masking detection of taxa, while strong nuisance covariance and artificial admixture risk to detecting pseudotaxa by mimicking categorical structures.
+Unfortunately, the use of taxometric analysis has remained limited in neurodevelopmental research, which most existing studies focusing on adult psychopathology [@beauchaine2003taxometrics; @haslam2020metaanalysis].
+In their debate on whether ASD is best conceptualized as a category or a continuum, @happe2021both and @chown2021bit do not even mention such methods.
+Similarly, in their influential review, @astle2022trandiagnostic briefly acknowledge taxometry, but puts much more emphasis on studies using cluster analysis.
+However, the limitations of clustering methods for assessing taxonicity have been discussed for decades: clustering algorithms tend to impose (rather than test for) categorical structures, they provide weak tests of taxonicity, and often fail to recover known taxa in simulations [@beauchaine2003taxometrics; @ruscio2004boundaryissues].
+Recent simulations using psychologically realistic data (featuring both skewness and intercorrelated indicators) suggest that commonly used clustering methods are prone to detecting pseudoclusters while presenting low power to identify true taxa [@toffalini2022entia; @toffalini2024clusters].
+Indeed, @astle2022trandiagnostic interpret clustering results as evidence that data-driven groups do not align with DSM-5 categories, rather than as compelling tests of latent structural form.
+
+Taxometric analysis is not without limitations, some of which are shared with cluster analysis.
+Commonly cited threats to validity of conclusions include poor indicator validity, nuisance covariance, low base rate, and artificial admixture [@cole2004taxometric; @haslam2020metaanalysis; @ruscio2004checklist; @ruscio2018using].
+
+Poor indicator validity implies that indicators present little mean difference between individuals with and without the target condition (i.e., the putative taxon and complement).
+Ideally, Cohen’s *d* for each indicator should substantially exceed 1 (*d* $\gg$ 1).
+Nuisance covariance refers to strong residual correlations among indicators after accounting for group membership (indicating redundancy among variables).
+Low base rate means that too few individuals in the sample belong to the taxon, making accurate detection difficult (samples with *n* < 30 for the taxonic group are usually deemed insufficient).
+This is particularly relevant when studying rare diagnostic categories.
+Artificial admixture occurs when participants are combined from different recruitment sources or selection criteria, often as an attempt to compensate for a low base rate.
+For instance, clinical cases may be recruited separately from controls and merged into a single sample.
+In general, poor indicator validity and low base rate reduce power to detect true taxa, while high nuisance covariance and artificial admixture increase the risk of detecting pseudotaxa by mimicking a categorical structure.
 
@@
 
-The main goal of this systematic review is to critically examine the use of taxometric methods in the study of neurodevelopmental disorders (NDDs) within the published literature.
-We focused our review on the most commonly studied NDD categories: Autism Spectrum Disorder (ASD), Attention-Deficit/Hyperactivity Disorder (ADHD), and all forms of Specific Learning Disorders (SLDs), along with the closely related condition of Language Disorder/Impairment (the latter was not included in the preregistered protocol and thus constitutes a deviation).
-We excluded Intellectual Disability, as its etiological basis is often well identified and taxonic by definition in many cases (e.g., Down syndrome), or associated with identifiable medical risk factors.
-Residual non-syndromic cases, in turn, are thought to reflect the lower tail of general mental ability, and thus as inherently dimensional.
+The primary aim of this systematic review is to critically examine the use of taxometric methods in studies of NDDs within the published literature.
+We focused on the most commonly investigated NDD categories: Autism Spectrum Disorder (ASD), Attention-Deficit/Hyperactivity Disorder (ADHD), and all forms of Specific Learning Disorders (SLDs), along with the closely related condition of Language Disorder/Impairment.
+(Language disorders were not included in the preregistered protocol and therefore constitute a minor deviation.) We excluded Intellectual Disability, as its etiology is often clearly identifiable and taxonic/syndromic by definition in many cases (e.g., Down syndrome), or linked to specific medical risk factors, while residual non-syndromic cases are typically regarded as reflecting the lower tail of general mental ability and thus inherently dimensional.
 
-A core aim of the review is to critically assess the methodological quality and credibility of findings reported in published taxometric studies.
-We evaluated potential sources of bias such as artificial admixture, and whether sufficiently complete methodological information was reported to judge other threats to validity, including low indicator validity, nuisance covariance, and the skewness of indicators.
-(The latter is relevant when scales assess rare or extreme behaviors, and can create the illusion of categorical structure).
+A crucial goal was to assess the methodological quality and credibility of taxometric findings.
+We specifically evaluated potential sources of bias, such as artificial admixture, and assessed whether studies provided sufficient methodological information to evaluate other known threats to validity, including low indicator validity, high nuisance covariance, and skewed indicator distributions.
+(The latter is particularly relevant when scales assess rare or extreme behaviors, which can create the illusion of a categorical structure.)
 
-Following this systematic review, we present a Monte Carlo simulation study to empirically demonstrate the extent to which artificial admixture can lead to false detection of categorical latent structures in dimensional data, both when using taxometric analysis and cluster analysis.
+Following the review, we report a Monte Carlo simulation study designed to empirically demonstrate the extent to which artificial admixture can lead to the spurious detection of categorical latent structures in otherwise dimensional data, using both taxometric and clustering methods.
 
@@
 
-The systematic review was conducted in accordance with the above listed goals, the preregistration protocol, and the PRISMA guidelines. The methods are detailed below.
+The systematic review was conducted in accordance with the Preferred Reporting items for Systematic Reviews and Meta-Analysis (PRISMA) 2020 guidelines.
+A review protocol was preregistered on OSF registries at the following link: https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0 The methods are detailed below.
 
@@
 
-A systematic search was conducted on four databases: PsycINFO, PubMed, Scopus, Web of Science.
-No time limits were set, however virtually no records from before 1990 were expected, as taxometric methods were not yet developed or diffuse then.
-The following search query was performed on titles, abstracts, and/or keywords:
+A systematic search for relevant peer-reviewed articles was conducted in April 2025 using four electronic databases: PsycINFO, PubMed, Scopus, and Web of Science.
+No time limits were applied, although records predating 1990 were not expected, as taxometric methods were not yet widely adopted at that time.
+The following search query—reflecting the methods and target populations of interest—was applied to titles, abstracts, and keywords:
 
-((“taxometric\*” OR “MAMBAC” OR “MAXCOV” OR “MAXEIG” OR “L-Mode”) AND (“autis\*” OR “ASD” OR “attention-deficit\*” OR “ADHD” OR “learning dis\*” OR “reading dis\*” OR “dyslexia” OR “math\* dis\*" OR "math\* diff\*"))
+```
+(("taxometric*" OR "MAMBAC" OR "MAXCOV" OR "MAXEIG" OR "L-Mode") AND 
+("autis*" OR "ASD" OR "attention-deficit*" OR "ADHD" OR "learning dis*" 
+OR "reading dis*" OR "dyslexia" OR "math* dis*" OR "math* diff*"))
+```
 
-Where the wildcard "\*" was not supported (i.e., PubMed) it was replaced with full terms ("disorder", "disability" instead of "dis\*"; "mathematical", "mathematics", "math" instead of "math\*").
+In databases where the wildcard `*` was not supported (e.g., PubMed), it was replaced with full terms (e.g., `"disorder"`, `"disability"` instead of `"dis*"`; `"mathematical"`, `"mathematics"`, `"math"` instead of `"math*"`).
 
-Language disorder was subsequently added as a deviation from preregistration, replacing the second part of the above search query with ("language imp\*" OR "language dis\*").
-In addition, the full search query was launched on Google Scholar and the first 10 pages were screened; the references of and citations to @haslam2020metaanalysis were searched on Google Scholar for additional results; and a deep research via OpenAI's gpt-4o was launched providing it the preregistration protocol as context.
-All promising records thus emerging and not previously identified via databases, were listed under the *"Identification of studies via other methods"* section of the PRISMA flowchart.
+Language disorder was subsequently added as a deviation from the preregistered protocol, replacing the second part of the search query with:
+
+```
+("language imp*" OR "language dis*")
+```
+
+Additionally, the full search query was launched on Google Scholar, and the first 10 pages of results were screened manually.
+References cited in, and articles citing, @haslam2020metaanalysis were also screened via Google Scholar.
+Finally, an exploratory search using OpenAI’s GPT-4o was conducted, using the preregistration protocol as contextual input.
+All promising records identified through these additional methods and not retrieved via database searches were listed under the *"Identification of studies via other methods"* section of the PRISMA flowchart.
 
@@
 
-Inclusion criteria:
+Inclusion criteria were as follows:  
+1. The study applied at least one taxometric procedure (e.g., MAMBAC, MAXCOV, L-Mode, MAXEIG);  
+2. The target population consisted of individuals with NDDs or related traits;  
+3. The study drew a conclusion about the latent structure (categorical vs. dimensional), or reported sufficient information (e.g., CCFIs) for an independent reader to infer such a conclusion.
 
--   The study applies at least one taxometric procedure (e.g., MAMBAC, MAXCOV, L-Mode, MAXEIG);
+Exclusion criteria were:  
+1. The article was theoretical, methodological, or a simulation study without empirical data;  
+2.
+The study did not report any behavioral data (e.g., only neuroimaging or neurophysiological data; note: this criterion was later reconsidered, but no studies were excluded for this reason in practice);
+3. The study was a review and/or meta-analysis;  
+4. The study employed latent structure methods other than taxometric analysis (e.g., latent profile analysis, mixture modeling), with no taxometric procedures reported.
 
--   The target population is individuals with NDDs or related traits;
-
--   The study draws a conclusion about the latent structure (categorical vs.
-dimensional) or reports sufficient information for an independent reader to reach a conclusion based on those results (e.g., reports CCFIs).
-
-Exclusion criteria:
-
--   Theoretical, methodological, or simulation articles without empirical data;
-
--   Articles not reporting any behavioral data (e.g., only neuroimaging or neurophysiological data);
-
--   Reviews and/or meta-analyses;
-
--   Studies using other latent structure methods only (e.g., LPA, mixture modeling) without any taxometric methods being employed.
+In addition, consistent with the preregistered protocol, the search was limited to published, peer-reviewed journal articles.
+This decision reflected the primary aim of providing a critical review of the existing empirical literature.
+However, one potentially relevant non-journal record was identified and, while excluded from the systematic review, is discussed separately in the *Discussion* section.
 
@@
 Two independent human coders (MC and RP) screened and coded all articles.
-All discrepancies were resolved through discussion, and later revised by the PI (ET).
-No AI tools were used in the process, except in occasional cases by the human coders to assist them clarify the content of articles.
+No AI tools were involved in the coding process, except in occasional cases where coders used them to help clarify specific article content.
+All discrepancies were resolved through discussion and later reviewed by the principal investigator (ET).
 
-Since a single article could report multiple (sub)samples and/or multiple taxometric analyses on the same sample, the full coded dataframe followed a long-form data structure, with some articles repeating over multiple rows.
-Coded variables are listed in the data dictionary table reported in the preregistration protocol, to which a few variables for notes and coders agreement were added.
+Since a single article could report multiple (sub)samples and/or multiple taxometric analyses on the same sample, the full coded dataset followed a long-form data structure, with some articles appearing across multiple rows.
+Coded variables were defined in the data dictionary included in the preregistration protocol, with a few additional fields added to capture coder notes and agreement metrics.
+
+Inter-coder agreement was fair to good across all phases.
+In the title and abstract screening phase, Cohen’s *κ* = 0.79 (34 out of 38 independent decisions matched).
+In the full-text eligibility screening, Cohen’s *κ* = 0.52 (15 out of 19).
+For the final data extraction, 84% of rows (71 out of 85) were extracted by both coders.
+Agreement on critical variables was high: for the "authors' taxonic conclusion" agreement was 94%; for "artificial admixture", it was 95% at the sample or subsample level.
 
@@
 
+A total of 110 records were identified through database searches.
+After removing 67 duplicates and excluding 5 additional records (dissertations or book chapters, only one of which appeared to include primary taxometric analyses but could not be retrieved), 38 unique articles were screened at the title and abstract level.
+
+Nineteen records were excluded at this stage.
+Full-text assessment was then conducted on the remaining 19 articles, of which 7 were excluded for not meeting the inclusion criteria: 4 did not apply taxometric methods, and 3 did not target neurodevelopmental disorders.
+
+An additional 4 records were identified through citation tracking and online browsing. Of these, 2 were excluded (1 was a methodological article, and 1 did not employ taxometric analysis).
+
+In total, 14 studies met inclusion criteria and were included in the review. The PRISMA flowchart in @fig-prisma summarizes the study selection process.
+
+@fig-prisma *about here*
+
+![PRISMA 2020 flow diagram illustrating the study screening and selection process](Figures/PRISMA_flow_diagram.png){#fig-prisma}
+
+```{r}
+df = data.frame(read_excel("../Literature-review/data/data_extraction.xlsx"))
+```
+
+A summary of the included studies, simplified to one row per article, is presented in @tbl-summary.
+A full reporting of all coded data for each individual taxometric analysis is available in the Supplemental Online Material.
+The total number of taxometric analyses was `r nrow(df)`, ranging from `r min(table(df$ID_article))` to `r max(table(df$ID_article))` analyses per article.
+The targeted disorders were ADHD (6 articles), ASD (6 articles), dyslexia (1 article), and Specific Language Impairment (SLI; 2 articles), with one article addressing both ADHD and ASD.
+Most articles (10 out of 14) examined whether a given disorder constituted a taxon, while the remaining 4 focused on whether putative subtypes within a disorder exhibited taxonic structure.
+
+The most commonly used taxometric method was MAMBAC, applied in 13 out of 14 articles and in `r sum(grepl("MAMBAC", df$Taxometric_methods_used))` out of `r nrow(df)` total analyses.
+This was followed by MAXEIG (9 articles), L-Mode (7 articles), MAXCOV (2 articles), and MAXSLOPE (1 article).
+Most studies employed more than one taxometric procedure, with a median of `r median(str_count(df$Taxometric_methods_used, ";") + 1)` methods per analysis.
+In 11 out of 14 articles, interpretation was based on a quantitative fit index (Fit~d~ or, in more recent studies, the CCFI), while the remaining 3 relied on visual inspection of taxometric curves.
+The number of indicators entered into analyses ranged from `r min(df$Indicators_number, na.rm = TRUE)` to `r max(df$Indicators_number, na.rm = TRUE)`, with a median of `r median(df$Indicators_number, na.rm = TRUE)` (excluding one article in which the number was not clearly reported).
+The median sample size across all analyses was `r median(df$Sample_size)`, or `r median((df |> group_by(ID_article) |> summarize(avg = mean(Sample_size)))$avg)` when averaged within articles.
+
+Taxonic conclusions were mixed.
+All five articles targeting ADHD as a disorder concluded in favor of a dimensional structure, as did both articles focusing on SLI.
+In contrast, all four articles examining ASD as a disorder leaned toward, or explicitly supported, a categorical (taxonic) structure.
+The remaining four articles investigated subtypes of disorders, and all leaned toward taxonic conclusions—either strongly or predominantly [@stevens2018functional on ADHD subtypes; @munson2008asd on ASD subtypes; @obrien2012dyslexia on dyslexia subtypes], or partially [@ingram2008defining, which supported ASD subtypes as taxa depending on the set of indicators used].
+In total, 8 out of 14 articles reached a taxonic conclusion in at least some of their taxometric analyses.
+Specifically, `r sum(df$Taxonic_conclusion)` out of `r nrow(df)` total analyses—representing `r round(mean(df$Taxonic_conclusion) * 100, 1)`%—yielded taxonic results.
+These were primarily associated with ASD: `r sum(grepl("ASD", df$Target_disorder) & df$Taxonic_conclusion == 1)` out of `r sum(grepl("ASD", df$Target_disorder))` analyses involving ASD (i.e., `r round(100 * sum(grepl("ASD", df$Target_disorder) & df$Taxonic_conclusion == 1) / sum(grepl("ASD", df$Target_disorder)), 1)`%) supported a categorical structure.
+
+Artificial admixture was evident or strongly suspected in at least part of the analyses in 8 out of 14 articles, involving a total of `r sum(df$Artificial_admixture)` out of `r nrow(df)` analyses.
+Notably, articles that incurred artificial admixture tended to align with those that reached taxonic conclusions: 6 of the 8 articles with suspected admixture also reported taxonic findings.
+This association was even more pronounced at the level of individual analyses.
+A taxonic conclusion was reached in `r sum(df$Taxonic_conclusion[df$Artificial_admixture == 1])` out of `r sum(df$Artificial_admixture == 1)` analyses (`r round(100 * mean(df$Taxonic_conclusion[df$Artificial_admixture == 1]), 1)`%) where admixture was evident or suspected.
+In contrast, only `r sum(df$Taxonic_conclusion[df$Artificial_admixture == 0])` out of `r sum(df$Artificial_admixture == 0)` analyses (`r round(100 * mean(df$Taxonic_conclusion[df$Artificial_admixture == 0]), 1)`%) reached a taxonic conclusion when admixture was likely or certainly absent.
+
+Remaining methodological issues concerned missing information about key descriptive statistics and validity indices.
+In particular, **nuisance covariance**—that is, correlations among indicators within diagnostic groups or within the putative taxon and complement—could be coded for only 2 out of 14 articles.
+Where reported, it was small to negligible in @james2016latent, but exceeded 0.40 for many indicators in @frazier2023asd; notably, both studies reached a taxonic conclusion.
+**Skewness** of indicator score distributions was reported in only 5 articles.
+When available, skewness values were often near or beyond ±1 for at least one indicator in roughly one-third of analyses.
+However, there was no clear association between skewness and whether a taxonic conclusion was reached.
+By contrast, **indicator validity** was more consistently reported, and could be coded in all but 3 articles.
+Where available, validity (typically expressed as Cohen’s *d*) was almost always large or very large (*d* $\gg$ 1), consistent with best-practice recommendations for taxometric analysis.
+
+@tbl-summary *about here*
+
+```{r}
+#| label: tbl-summary
+#| tbl-cap: Overview summary of the reviewed articles."
+summary_tab = read_excel("Tables/Review-Summary.xlsx",sheet=2)
+kable(summary_tab, format = "latex", booktabs = TRUE) |>
+  kable_styling(font_size = 8, latex_options = c("striped", "scale_down")) |>
+  column_spec(1:ncol(summary_tab),width="5em") |>
+  landscape()
+```
+
+## Additional Use of Cluster Analysis
+
+Among the 14 identified articles, 6 also employed clustering methods—typically latent class analysis (LCA) or latent profile analysis (LPA).
+In 5 of these 6 cases, clustering was conducted in parallel with taxometric analysis [@james2016latent; @munson2008asd; @frazier2023asd; @frazier2007adhd; @deserno2023probing], while in one case it was presented as a secondary analysis [@frazier2010asd].
+All six studies targeted ADHD and/or ASD, and in all cases clustering results were interpreted as supporting taxonic conclusions.
+These interpretations generally converged with those from taxometric analysis, with one exception: in @frazier2007adhd, clustering results were partially inconclusive but suggested the presence of subgroups within ADHD, thus contrasting with the dimensional structure favored by the taxometric analysis.
+In all but two studies [@frazier2007adhd; @munson2008asd], we identified at least some risk of artificial admixture.
+
 # Simulation of Artificial Admixture
 
-We focus on artificial admixture because it is a very powerful threat to validity that could relatively easily be avoided by design when planning a study, and because as suggested by the result of the systematic review presented below, it is unfortunately frequent.
-There are different types of admixtures; @ruscio2004checklist discusses three cases.
-First, as mentioned above, the situation where clinical and nonclinical cases are separately recruited and merged.
-Second, the case of splitting a sample into subsamples (using alternative criteria) and then using those to test different taxometric questions.
-Third, trimming away observations of the putative complement (i.e., individuals without the disorder) to increase the base rate.
-All of these procedures artificially impose categorical structures to the data that do not exist in the population.
-It is worth mentioning that artificial admixture is also a threat for any type of cluster analysis, for the exact same reasons that it is a problem for taxometric analysis.
+We focus on artificial admixture because it represents a powerful threat to validity that is both common (as suggested by our review) and avoidable by design through better planning of a study.
+Different types of admixture can be incurred.
+@ruscio2004checklist describes three main forms.
+First (and most frequently) is the case where clinical and nonclinical participants are separately recruited and merged into a single sample.
+Second, admixture can occur when a sample is artificially split into subsamples based on different selection criteria, and these are then used to test different taxometric questions [cf.
+"subtractive" compound sampling in @haslam2020metaanalysis].
+Third, researchers may trim observations from the putative complement (i.e., individuals without the disorder) to inflate the base rate of the taxon.
+Each of these procedures risks imposing a categorical structure onto data that might be dimensional in nature.
+It is worth noting that artificial admixture poses a threat not only to taxometric analysis, but also to clustering methods, for the same reasons: both infer latent structure based on observed covariation, which can be distorted by selection artifacts.
 
-We simulate a dimensional scenario that makes it visually compelling how artificial admixture of the first type mentioned above creates pseudotaxa.
-A single normally-distributed latent variable $X$ is generated, that defines the target condition.
-A series of observed indicators $x1, x2, etc.$ are simulated, incorporating the $X$ latent variable plus some random normally-distributed noise ($R^2 = 0.77$).
-Individuals above the 99^th^ percentile of $X$ receive a diagnosis via clinical assessment.
-A sample of $N = 1000$ individuals is collected from the general population.
-Since the base rate is low (on average, only about $n = 10$ cases are expected to naturally present a diagnosis), researchers select and recruit an additional $n = 1000$ cases with a clinical diagnosis from the general population, thus creating artificial admixture.
-All data simulation is available in the online materials at xxxxxxxxxxxxxxxx.
+```{r}
+load("../R/simulateAdmixtResults.RData")
+```
 
-An instance of the simulation, plotting two observed variables for illustrative purposes, is shown in @FIGURE_xx.
-Color is added for further clarity to distinguish individuals with and without a diagnosis, but two clusters still clearly emerge.
-Indicator validity is good (about $Cohen's$ $d = 2$), while nuisance covariance is partly problematic: for the clinical cases $r \approx 0$, but for the nonclinical cases $r = 0.61$.
-It is worth reminding that no taxa exist here, and the appearance of a categorical structure is mainly imposed by admixture, as shown below.
+We simulate a dimensional scenario to visually demonstrate how artificial admixture, specifically of the first type described above, can produce pseudotaxonic results.
+A single normally distributed latent variable $X$ is generated to represent the underlying clinical trait of interest.
+A set of observed indicators ($x_1$, $x_2$, $x3$) are then simulated as linear functions of $X$ plus normally distributed random noise, yielding an $R^2$ of 0.77 for each indicator.
+Individuals above the 99th percentile of $X$ are assumed to receive a diagnosis via global clinical assessment.
+A sample of $N = 1000$ individuals is drawn from the general population.
+Given the low base rate, only about $n = 10$ cases are expected to naturally receive a diagnosis.
+To address this, researchers draw a second sample consisting of an additional $n = 1000$, targeting clinically diagnosed cases, and merge it with the original sample, thereby creating an artificial admixture.
+The full simulation code and data are available in the online materials:
+<https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba>
 
-A Monte Carlo simulation was then performed, with 1000 iterations.
-Taxometric analysis was conducted using the "RTaxometric" package of R.
-As recommended, CCFI was used for inference: CCFI varies between 0 (maximal evidence for dimensional structure) to 1 (maximal evidence for categorical structure), with 0.5 indicating inconclusive results.
-Performed procedures were, as per default: MAMBAC, MAXEIG, and L-Mode.
-A third $x3$ observed indicator (similar to those above) was entered to allow conducting all these taxometric procedures.
-Over 1000 simulated datasets, results always strongly favored a categorical structure: average combined CCFI = xxxx, 95% quantile interval \[xxxx, xxxx\] (average MAMBAC CCFI = xxxx; average MAXEIG CCFI= xxxx; average L-Mode CCFI = xxxx).
-The smallest combined CCFI over 1000 iterations was xxxx.
-To confirm that artificial admixture also leads to misleading results when performing clustering, Gaussian Mixture Models (GMM) was also run testing 1 vs 2 components, using the "mclust" package of R: over 1000 iterations, results always favored the 2-component relative to the 1-component model, with an average $\Delta BIC = xxxx$ in favor of the 2-component model.
+An instance of the simulation, illustrating two observed variables $x_1$ and $x_2$, is presented in @fig-admixture.
+Color is used to distinguish individuals with and without a diagnosis, but even without this aid, two clusters visually emerge.
+This apparent separation is misleading: no latent categories exist in the data.
+The appearance of a categorical structure is a consequence of artificial admixture.
+In this simulated example, indicator validity is high (Cohen’s *d* ≈ 2), suggesting strong separation between groups on each variable.
+However, nuisance covariance is partly problematic: among diagnosed individuals, the correlation between indicators is close to zero ($r \approx 0.05$), whereas among non-diagnosed individuals, it is substantial ($r \approx 0.60$).
+This asymmetry, combined with differential recruitment, contributes to the illusion of a latent taxon.
 
-To confirm that the categorical structure was incorrectly favored specifically because of artificial admixture, we run another Monte Carlo simulation like the one above, but without admixture.
+@fig-admixture *about here*
+
+![A simulated example of artificial admixture, where a large number of individuals with a diagnosis are separately recruited and added to the sample (total N = 2000).
+The true underlying structure is strictly dimensional.](Figures/figureAdmixture.png){#fig-admixture}
+
+A Monte Carlo simulation was conducted with 1000 iterations to assess the impact of artificial admixture on taxometric outcomes.
+Analyses were performed using the `RTaxometric` package in R.
+As recommended in the literature, inference was based on the Comparison Curve Fit Index (CCFI), which ranges from 0 (indicating strong support for a dimensional structure) to 1 (indicating strong support for a categorical structure), with values around 0.5 considered ambiguous.
+The default procedures (MAMBAC, MAXEIG, and L-Mode) were applied to each simulated dataset.
+To enable all three procedures, three observed indicators ($x_1$, $x_2$, $x_3$) were included in the simulations.
+
+Over 1000 simulated datasets, results consistently and strongly favored a categorical interpretation: median combined CCFI = `r formatC(median(resultsAdmix$CCFI.mean),2,format="f")`, 95% quantile interval \[`r formatC(quantile(resultsAdmix$CCFI.mean,.025),2,format="f")`, `r formatC(quantile(resultsAdmix$CCFI.mean,.975),2,format="f")`\]; median MAMBAC CCFI = `r formatC(median(resultsAdmix$CCFI.MAMBAC),2,format="f")`; average MAXEIG CCFI= `r formatC(median(resultsAdmix$CCFI.MAXEIG),2,format="f")`; average L-Mode CCFI = `r formatC(median(resultsAdmix$CCFI.LMode),2,format="f")`.
+The smallest combined CCFI over 1000 iterations was `r formatC(min(resultsAdmix$CCFI.mean),2,format="f")`.
+Therefore, admixture can be an extremely strong driver of false categorical findings, especially when many clinical cases are separately recruited and added to a sample.
+
+To verify that artificial admixture also produces misleading results when using clustering methods, we ran Gaussian Mixture Models (GMMs) comparing 1-component versus 2-component solutions, using the `mclust` package in R.
+Model selection was based on the Bayesian Information Criterion (BIC), as per default in GMMs.
+GMM is a form of model-based clustering grounded in finite mixture modeling.
+It is conceptually similar to latent profile analysis (LPA; with which it is often confused) but is more flexible in that it models covariances among indicators.
+Across 1,000 simulated datasets, results consistently favored the 2-component solution over the unimodal alternative.
+The median difference in BIC (ΔBIC) was `r formatC(median(resultsAdmix$gmmBIC1 - resultsAdmix$gmmBIC2), 2, format = "f")`, in favor of the 2-component model.
+
+To confirm that the categorical structure was incorrectly favored specifically due to artificial admixture, we run a second Monte Carlo simulation using the same dimensional data-generating process as before, but without admixture.
 In this case, a single sample of $N = 2000$ observations was drawn at each iteration.
-Therefore, nuisance covariance remains strong, but there is no admixture.
-Average combined CCFI = xxxx, 95% quantile interval \[xxxx, xxxx\]; average MAMBAC CCFI = xxxx; average MAXEIG CCFI= xxxx; average L-Mode CCFI = xxxx.
-Largest combined CCFI was xxxx.
-Thus, results remain frequently around inconclusive, probably due to nuisance covariance, but never strongly favor a categorical structure.
-GMM also consistently favored the 1-component model (over the 2-component model), with an average $\Delta BIC = xxxx$ in favor of the 1-component model.
-Therefore, admixture can be an extremely strong driver of false categorical findings, especially when many clinical cases are separately recruited and added to the sample.
+Therefore, nuisance covariance remains present, but there is no admixture.
+Median combined CCFI = `r formatC(median(resultsNoAdmix$CCFI.mean),2,format="f")`, 95% quantile interval \[`r formatC(quantile(resultsNoAdmix$CCFI.mean,.025),2,format="f")`, `r formatC(quantile(resultsNoAdmix$CCFI.mean,.975),2,format="f")`\]; median MAMBAC CCFI = `r formatC(median(resultsNoAdmix$CCFI.MAMBAC),2,format="f")`; average MAXEIG CCFI= `r formatC(median(resultsNoAdmix$CCFI.MAXEIG),2,format="f")`; average L-Mode CCFI = `r formatC(median(resultsNoAdmix$CCFI.LMode),2,format="f")`.
+Clustering with GMM also consistently favored the 1-component model (over the 2-component model), with an average ΔBIC = `r formatC(median(resultsNoAdmix$gmmBIC2-resultsNoAdmix$gmmBIC1),2,format="f")` in favor of the 1-component model.
+
+Thus, without admixture taxometric results leaned towards a dimensional conclusion, but frequently remained inconclusive.
+This is probably due to nuisance covariance.
+It is worth noting, however, that under a dimensional structure achieving both high indicator validity and low within-group covariance might be practically impossible.
+This is because indicators that clearly differentiate individuals with and without a diagnosis must be strongly associated with the underlying trait, an association that produces inter-indicator covariance.
+In fact, this trade-off may iteself serve as indirect evidence for dimensionality.
 
@@
 
-The present review suggests that taxometric analysis has been underused in neurodevelopmental research, despite its unique relevance for the recent debate on the categorical vs dimensional nature of disorders [@astle2022trandiagnostic; @michelini2024world; @happe2021both].
-Focusing on selected NNDs, we identified 14 relevant records.
-The underuse of taxometric analysis in this field was expected: @haslam2020metaanalysis identified 183 articles employing taxometric methods on psychopathology, of which only 15 concerned childhood (these included various NNDs but also other conditions such as sleep problems).
-What is most worrying, many of the articles that we reviewed present serious methodological limitations and/or incomplete reporting of relevant information.
+We found that taxometric analysis remains underused in neurodevelopmental research, which is unfortunate given the unique relevance of this method to the ongoing debate about the nature of NDDs [@astle2022trandiagnostic; @michelini2024world; @happe2021both; @sonuga2020kuhnian].
+That underuse was expected, however.
+@haslam2020metaanalysis classified only 15 out of 183 taxometric studies in the "childhood" category (which included heterogeneous conditions, and even sleep problems).
+In the studies identified in our review, only ASD, ADHD, and SLI were targeted as disorders in taxometric analysis.
+None of learning disorders was targeted as a taxon, with Dyslexia being covered in one single article, but only for investigating its subtypes.
 
-In the simulated example, artificial admixture was also accompanied by large nuisance covariance, although it was demonstrated that the latter posed a less severe problem to inference.
-In fact, the coexistence of low nuisance covariance and good indicator validity is practically impossible under a dimensional scenario.
-This is because an indicator can largely separate individuals with vs without a diagnosis if it is strongly related to the latent dimension, but this leads to nuisance covariance across valid indicators.
-Ultimately, however, such impasse might be interpreted as a direct signal that the true latent structure is dimensional.
+Taxometric results appeared relatively consistent.
+Dimensional conclusions were predominantly reached for ADHD and SLI, while categorical conclusions prevailed for ASD and studies examining subtypes of any disorder.
+This echoes @haslam2020metaanalysis, and may seem to offer a compelling answer to the debate raised by @chown2021bit and @happe2021both.
+However, a closer inspection of the methodological limitations in the reviewed studies suggests that such conclusions should be interpreted with caution.
 
-While the DSM-5 may appear to promote a categorical model of disorders, it does not explicitly commit to a taxonic view.
-It is often researchers, not diagnostic manuals, who reify clinical categories as natural kinds, despite their originally pragmatic intent.
+We identified four studies that targeted ASD as a disorder in taxometric analyses, and all of them reached taxonic conclusions.
+However, some artificial admixture was identified in all of them, and it represents a major threat to conclusion validity, as demonstrated both in our data simulation and in prior literature [@ruscio2004checklist; @ruscio2004boundaryissues].
+Admixture in the reviewed studies ranged from sampling participants through different methods, only some of which explicitly targeted clinical populations [@james2016latent], to combining clinical and non-clinical samples [@frazier2023asd; @deserno2023probing].
+In one case, inclusion criteria from a larger database were partly unclear, but the sample was explicitly structured by diagnostic status [@deserno2023probing].
+Another form of admixture involved constructing a sample of affected and unaffected siblings [@frazier2010asd].
+According to our interpretation, the only relatively unbiased taxometric evidence suggesting that ASD may be a taxon as a disorder comes from a secondary analysis by @frazier2010asd, in which only unaffected siblings were examined.
 
-The fact that observed scores in psychology often deviate from normality [@micceri1989unicorn] may reflect psychometric characteristics (e.g., skewed indicators when observing times, accuracies, bounded sum scores) rather than the true shape of the underlying latent traits.
+Concerning subtypes, the situation was similar.
+All four reviewed articles reached some taxonic conclusions, but three of them likely involved some form of artificial admixture.
+However, the issue appeared to be nuanced and less severe in this subset of studies.
+For example, @obrien2012dyslexia used psychometric cutoffs across different tools and inclusion criteria, which may not necessarily have biased the results unless the selection criteria correlate with different underlying cognitive impairments (phonological vs.
+non-phonological).
+However, the authors themselves acknowledged this possibility, writing that *"depending on the dyslexia criterion used, samples may vary in composition with regard to their subtypes"* (p.33).
+
+Overall, the presence of artificial admixture was most frequently associated with taxonic conclusions.
+Only two notable mismatches were identified: @marcus2012latent reached a dimensional conclusion regarding ADHD despite having incurred admixture of clinical cases, while @munson2008asd reached a taxonic conclusion about ASD subtypes without clear evidence of admixture (all children had received an ASD diagnosis based on what appeared to be a uniform clinical assessment).
+As discussed above, a third exception was observed in a secondary analysis of a subsample by @frazier2010asd targeting ASD.
+Taken together, these findings reinforce the view that ADHD is best conceptualized as a dimensional condition, while ASD may exhibit taxonic features, although the latter conclusion is supported by limited unbiased evidence.
+
+In our simulated example, we found that artificial admixture represents a major threat to the validity of taxometric analysis, and to clustering methods as well.
+One might argue that our simulated scenario, in which half of the diagnosed cases were separately sampled, was extreme.
+However, this design is broadly consistent with the degree of admixture observed in the reviewed literature.
+It is worth noting that admixture is not limited to the obvious case of oversampling diagnosed individuals, and sometimes occurs in subtle ways.
+For example, through the undersampling of individuals with low trait levels, or by selectively drawing subsamples based on alternative clinical criteria within a larger dataset, as is often the case in subtype studies [cf.
+"subtractive" compound sampling, @haslam2020metaanalysis].
+
+An important issue related to the present discussion concerns how we should treat traditional diagnostic categories such as those found in the DSM-5, which have recently come under criticism [@astle2022trandiagnostic; @sonuga2020kuhnian; @sonuga2022dontfit].
+We argue that the question of whether clinically relevant conditions reflect discrete categories or multivariate, correlated continua does not necessarily undermine the practical utility of diagnostic systems.
+While the DSM-5 may seem to endorse a categorical model, it remains atheoretical and does not explicitly or scientifically commit to a taxonic view.
+Researchers, rather than diagnostic manuals, may risk reifying these categories as natural taxa, thus overinterpreting their conceptual status, hence we advocate for more high-quality taxometric studies.
+
+## Limitations and Future Directions
+
+A first and obvious limitation of the present review concerns the small number of studies identified.
+This prevented us from conducting a meta-analysis: the available evidence was too scattered, methodologically heterogeneous, and inappropriate for quantitative synthesis.
+However, this limitation is itself informative, underscoring the need for more taxometric investigation in the field of NDDs.
+While the search query was comprehensive and no time limits were applied, more studies might have been included by relaxing the eligibility criteria.
+However, this is unlikely to have made a substantial difference.
+Exclusion criteria specified the lack of use of behavioral indicators, but no study was excluded on this basis.
+One record was screened out for being a dissertation [@clemons2006taxometric].
+The full document could not be retrieved.
+Based on its abstract, it is clear that it focused on whether ADHD subtypes represent distinct taxa, and it reached a taxonic conclusion.
+However, some degree of admixture was possible (participants were separately recruited from clinical centers and schools).
+
+A relevant future direction for this line of investigation concerns a systematic examination of the use and results of cluster analysis.
+As described above, some of the reviewed articles employed clustering methods, and clustering is a prominent tool in this field [@astle2022trandiagnostic].
+We found that model-based approaches, such as latent class/profile analysis, were the most frequently used.
+The use of clustering is not without problems.
+First, while it is more flexible and can address broader questions, it does not offer a direct or consistent test of taxonicity [@beauchaine2003taxometrics; @ruscio2004boundaryissues].
+Second, meaningful scientific inference from clustering (beyond its simple use as a data reduction technique) requires careful handling of assumptions, as the risk of detecting pseudoclusters is high [@toffalini2022entia; @toffalini2024clusters].
+Third, it is is not immune to risks such as artificial admixture.
+Nonetheless, clustering and taxometric analysis could be more directly, but also critically, examined and compared when addressing the question of whether neurodevelopmental conditions reflect categories or dimensions.
 

### 2025-07-25 Paper/Paper.qmd
update paper

@@
 Testing the dimensional hypothesis in neurodevelopmental disorders", funded by Next Generation EU, Mission 4, Componente 1, CUP C53D23004210006'
-    gratitude: "We are grateful to Athanasia Vasileiadou who first pointed us to some of the reviewed articles."
+    gratitude: "We are grateful to Athanasia Vasileiadou who first pointed us to the topic, and to some of the reviewed articles."
     data-sharing: https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba
@@
     authorship-agreements: null
-abstract: "Taxometric methods are designed to test whether latent structures underlying psychological constructs are categorical (taxonic) or dimensional.
-Despite their relevance to the nature of neurodevelopmental disorders (NDDs), their application has been limited and often methodologically flawed.
-We conducted a preregistered systematic review to critically evaluate this literature.
-The review identified all published taxometric studies on autism, ADHD, learning, and language disorders, with a focus on methodological biases such as artificial admixture.
-While studies on ADHD and SLI supported dimensional structures, most studies on autism and NDD subtypes reported taxonic findings, often in the presence of methodological threats.
-A Monte Carlo simulation illustrates how artificial admixture can spuriously generate taxonic results in otherwise dimensional data.
-We conclude that evidence for categorical models in NDDs is frequently undermined by methodological artifacts, reinforcing the need for caution, but also advocating for more, high-quality taxometric studies in the field of NDDs."
+abstract: "Taxometric analysis is explicitly designed to test whether the latent structures underlying clinical psychological constructs are categorical (taxonic) or dimensional.
+Despite its relevance to ongoing debates about the nature of neurodevelopmental disorders (NDDs), its application in this field has been limited.
+We conducted a preregistered systematic review to critically evaluate the literature, with a focus on methodological threats such as artificial admixture.
+All published taxometric studies on autism, ADHD, learning disorders, and language impairment were identified and assessed.
+Studies on ADHD and language impairment generally supported dimensional structures, whereas most studies on autism and NDD subtypes predominantly reported taxonic findings.
+However, artificial admixture was frequently observed and undermines the credibility of many taxonic conclusions.
+To illustrate this risk, we conducted a Monte Carlo simulation showing how admixture can spuriously generate pseudotaxonic results.
+We urge caution in interpreting the outcomes, and advocate for more high-quality taxometric studies in this area."
 keywords: [Neurodevelopmental disorders, Taxometric analysis, Autism, ADHD, Learning disorders]
@@
 Traditional diagnostic categories defined in the DSM-5 [@dsm5] have increasingly been questioned in favor of a dimensional and transdiagnostic perspective [e.g., @astle2022trandiagnostic; @michelini2024world; @happe2020looking; @happe2021both; @sonuga2020kuhnian; @caviola2024editorial].
-Notably, the DSM-5 itself has incorporated some dimensional features (for example by defining a broad Autism Spectrum Disorder [ASD] in place of distinct diagnoses).
-Several factors have motivated this shift: the marked heterogeneity within diagnostic groups, the weak correspondence between data-driven clusters and DSM-based categories, the frequent comorbidity and phenotypic correlations across conditions, and the view that individuals with NDDs may represent, at least in part, the tails of developmental continua, quantitatively rather than qualitatively different from the so-called neurotypical population [@michelini2024world; @happe2021both; @mammarella2021nocore].
+Notably, the DSM-5 itself has incorporated some dimensional features (for example by defining a broad Autism Spectrum Disorder \[ASD\] in place of distinct diagnoses).
+Several factors have motivated this shift: the marked heterogeneity within diagnostic groups, the weak correspondence between data-driven clusters and DSM-based categories, the frequent comorbidity and phenotypic correlations across conditions, and the view that individuals with NDDs may represent the tails of developmental continua, quantitatively rather than qualitatively different from the so-called neurotypical population [@michelini2024world; @happe2021both; @mammarella2021nocore].
 
-Determining whether NDDs truly represent categories, rather than just tails of continuous distributions spanning the general population, is important for several reasons, including questions of social and personal identity for affected individuals, implications for diagnostic framing, comorbidity, and intervention choices in clinical contexts, and consequences for the organization of classrooms, educational practices, and even the design and power analysis of research studies [@caviola2024editorial].
-But more fundamentally, this question is relevant for our understanding of the ontological status of psychological phenomena.
+Determining whether NDDs represent categories, rather than just tails of continuous distributions spanning the general population, is important for several reasons, including questions of social and personal identity for affected individuals, implications for diagnostic framing, comorbidity, and intervention choices in clinical contexts, and consequences for the organization of classrooms, educational practices, and even the design and power analysis of research studies [@caviola2024editorial].
+More fundamentally, this question is relevant for our understanding of the ontological status of psychological phenomena.
 
@@
 Most behavioral traits are understood to result not from one or a few causes, but from the additive effects of many small influences [@chabris2015fourth].
-The central limit theorem [@fischer2011central] posits that, when many individual causes (such as discrete genetic variants) contribute additively, the resulting distribution tends to continuity and normality (this is, in fact, the "secret" of the Gaussian distribution).
+According to the central limit theorem [@fischer2011central], when many individual causes (e.g., discrete genetic variants) contribute additively, the resulting trait tends to approximate a continuous, Gaussian distribution.
 Nonetheless, a principled categorical view still resists in neurodevelopmental research, for example concerning ASD [e.g., @happe2021both; @chown2021bit; @frazier2023asd], subtypes within NDDs [e.g., @burgess2018taxodyslexia] or the proposed emergence of data-driven taxa [@astle2022trandiagnostic].
 
-Taxometric analysis is a set of quantitative methods originally introduced by Paul Meehl and colleagues in the 1970s and refined over decades [e.g., @meehl1995taxometric], explicitly aimed at establishing whether a clinical disorder is characterized by an underlying categorical or dimensional structure.
+Taxometric analysis is a set of quantitative methods originally introduced by Paul Meehl and colleagues and refined over decades [@meehl1995taxometric], explicitly aimed at establishing whether a clinical disorder is characterized by an underlying categorical or dimensional structure.
 It has been widely applied in psychopathology, and typically suggested that most mental health conditions (e.g., depression, anxiety) are better conceptualized as dimensional [@haslam2012review; @haslam2020metaanalysis].
-In brief, taxometric procedures, such as MAMBAC, MAXCOV, MAXEIG, and L-Mode examine covariation patterns among observed indicators to compare competing latent models: one categorical, the other dimensional [@ruscio2007taxometric; @ruscio2004checklist; @cole2004taxometric].
-These methods now feature simulation-based tools like the Comparison Curve Fit Index (CCFI), which provide a quantitative measures to facilitate inference [@ruscio2018using].
+In brief, taxometric procedures, such as MAMBAC, MAXCOV, MAXEIG, and L-Mode examine covariation patterns among observed indicators to adjudicate between competing latent categorical vs dimensional models [@ruscio2007taxometric; @ruscio2004checklist; @cole2004taxometric].
+These methods now feature simulation-based tools like the Comparison Curve Fit Index (CCFI), which provide quantitative measures to facilitate inference [@ruscio2018using].
 
-Unfortunately, the use of taxometric analysis has remained limited in neurodevelopmental research, which most existing studies focusing on adult psychopathology [@beauchaine2003taxometrics; @haslam2020metaanalysis].
+Unfortunately, the use of taxometric analysis has remained limited in neurodevelopmental research, with most existing studies focusing on adult psychopathology [@beauchaine2003taxometrics; @haslam2020metaanalysis].
 In their debate on whether ASD is best conceptualized as a category or a continuum, @happe2021both and @chown2021bit do not even mention such methods.
-Similarly, in their influential review, @astle2022trandiagnostic briefly acknowledge taxometry, but puts much more emphasis on studies using cluster analysis.
+In their influential review, @astle2022trandiagnostic briefly acknowledge taxometry, but predominantly emphasize studies using cluster analysis.
 However, the limitations of clustering methods for assessing taxonicity have been discussed for decades: clustering algorithms tend to impose (rather than test for) categorical structures, they provide weak tests of taxonicity, and often fail to recover known taxa in simulations [@beauchaine2003taxometrics; @ruscio2004boundaryissues].
-Recent simulations using psychologically realistic data (featuring both skewness and intercorrelated indicators) suggest that commonly used clustering methods are prone to detecting pseudoclusters while presenting low power to identify true taxa [@toffalini2022entia; @toffalini2024clusters].
+Recent simulations using psychologically realistic data (featuring both some skewness and intercorrelated indicators) suggest that commonly used clustering methods are prone to detecting pseudoclusters while presenting low power to identify true taxa [@toffalini2022entia; @toffalini2024clusters].
 Indeed, @astle2022trandiagnostic interpret clustering results as evidence that data-driven groups do not align with DSM-5 categories, rather than as compelling tests of latent structural form.
@@
 Ideally, Cohen’s *d* for each indicator should substantially exceed 1 (*d* $\gg$ 1).
-Nuisance covariance refers to strong residual correlations among indicators after accounting for group membership (indicating redundancy among variables).
-Low base rate means that too few individuals in the sample belong to the taxon, making accurate detection difficult (samples with *n* < 30 for the taxonic group are usually deemed insufficient).
-This is particularly relevant when studying rare diagnostic categories.
+Nuisance covariance refers to strong residual correlations among indicators after accounting for group membership, indicating redundancy.
+Low base rate means that too few individuals in the sample belong to the taxon, limiting power (samples with *n* \< 30 for the taxonic group are usually deemed insufficient).
+This is particularly relevant when studying rare conditions.
 Artificial admixture occurs when participants are combined from different recruitment sources or selection criteria, often as an attempt to compensate for a low base rate.
@@
 The primary aim of this systematic review is to critically examine the use of taxometric methods in studies of NDDs within the published literature.
-We focused on the most commonly investigated NDD categories: Autism Spectrum Disorder (ASD), Attention-Deficit/Hyperactivity Disorder (ADHD), and all forms of Specific Learning Disorders (SLDs), along with the closely related condition of Language Disorder/Impairment.
-(Language disorders were not included in the preregistered protocol and therefore constitute a minor deviation.) We excluded Intellectual Disability, as its etiology is often clearly identifiable and taxonic/syndromic by definition in many cases (e.g., Down syndrome), or linked to specific medical risk factors, while residual non-syndromic cases are typically regarded as reflecting the lower tail of general mental ability and thus inherently dimensional.
+We focused on the most commonly investigated NDD categories: Autism Spectrum Disorder (ASD), Attention-Deficit/Hyperactivity Disorder (ADHD), Specific Learning Disorders (SLDs), and the closely related condition of Language Disorder/Impairment.
+(The latter was not included in the preregistered protocol and constitutes a minor deviation.) We excluded Intellectual Disability, as its etiology is often clearly identifiable and taxonic/syndromic by definition in many cases (e.g., Down syndrome), or linked to specific medical risk factors, while residual non-syndromic cases are typically regarded as reflecting the lower tail of general mental ability and thus inherently dimensional.
 
@@
 We specifically evaluated potential sources of bias, such as artificial admixture, and assessed whether studies provided sufficient methodological information to evaluate other known threats to validity, including low indicator validity, high nuisance covariance, and skewed indicator distributions.
-(The latter is particularly relevant when scales assess rare or extreme behaviors, which can create the illusion of a categorical structure.)
+(Strong skewness can create the illusion of a categorical structure, and it might be relevant when assessing rare or extreme behaviors.)
 
-Following the review, we report a Monte Carlo simulation study designed to empirically demonstrate the extent to which artificial admixture can lead to the spurious detection of categorical latent structures in otherwise dimensional data, using both taxometric and clustering methods.
+Following the review, we report a Monte Carlo simulation study designed to empirically illustrate the extent to which artificial admixture can lead to the spurious detection of categorical latent structures in otherwise dimensional data.
 
@@
 
-Inclusion criteria were as follows:  
-1. The study applied at least one taxometric procedure (e.g., MAMBAC, MAXCOV, L-Mode, MAXEIG);  
-2. The target population consisted of individuals with NDDs or related traits;  
-3. The study drew a conclusion about the latent structure (categorical vs. dimensional), or reported sufficient information (e.g., CCFIs) for an independent reader to infer such a conclusion.
+Inclusion criteria were as follows:\
+1. The study applied at least one taxometric procedure;\
+2. The target population consisted of individuals with NDDs or related traits;\
+3. The study drew a conclusion about the latent structure (categorical vs. dimensional), or reported sufficient information (e.g., CCFIs) to infer such a conclusion.
 
-Exclusion criteria were:  
-1. The article was theoretical, methodological, or a simulation study without empirical data;  
-2.
-The study did not report any behavioral data (e.g., only neuroimaging or neurophysiological data; note: this criterion was later reconsidered, but no studies were excluded for this reason in practice);
-3. The study was a review and/or meta-analysis;  
-4. The study employed latent structure methods other than taxometric analysis (e.g., latent profile analysis, mixture modeling), with no taxometric procedures reported.
+Exclusion criteria were:\
+1. The article was theoretical, methodological, or a simulation study without empirical data;\
+2. The study did not report any behavioral data (in practice, no screened study was excluded for this criterion);\
+3. The study was a review and/or meta-analysis;\
+4. The study employed latent structure methods other than taxometric analysis (e.g., only clustering), with no taxometric procedures reported.
 
-In addition, consistent with the preregistered protocol, the search was limited to published, peer-reviewed journal articles.
-This decision reflected the primary aim of providing a critical review of the existing empirical literature.
-However, one potentially relevant non-journal record was identified and, while excluded from the systematic review, is discussed separately in the *Discussion* section.
+In addition, following the preregistered protocol, the search was limited to published, peer-reviewed journal articles.
+This decision reflected the aim of providing a critical review of the existing empirical literature.
+Only one potentially relevant non-journal record was identified and, while excluded from the systematic review, is discussed separately in the *Limitation* section.
 
@@
 Two independent human coders (MC and RP) screened and coded all articles.
-No AI tools were involved in the coding process, except in occasional cases where coders used them to help clarify specific article content.
-All discrepancies were resolved through discussion and later reviewed by the principal investigator (ET).
+No AI tools conducted any screening or coding, although coders occasionally used them to help clarify specific article content.
+All discrepancies were resolved through discussion and were later reviewed by the PI (ET).
 
@@
 #| label: tbl-summary
-#| tbl-cap: Overview summary of the reviewed articles."
+#| tbl-cap: Overview summary of the reviewed articles.
 summary_tab = read_excel("Tables/Review-Summary.xlsx",sheet=2)
@@
 
-# Simulation of Artificial Admixture
+# A Monte Carlo Simulation of Artificial Admixture
 
-We focus on artificial admixture because it represents a powerful threat to validity that is both common (as suggested by our review) and avoidable by design through better planning of a study.
+Artificial admixture represents a powerful threat to validity that is common but also easily avoidable by design.
 Different types of admixture can be incurred.
-@ruscio2004checklist describes three main forms.
-First (and most frequently) is the case where clinical and nonclinical participants are separately recruited and merged into a single sample.
-Second, admixture can occur when a sample is artificially split into subsamples based on different selection criteria, and these are then used to test different taxometric questions [cf.
-"subtractive" compound sampling in @haslam2020metaanalysis].
-Third, researchers may trim observations from the putative complement (i.e., individuals without the disorder) to inflate the base rate of the taxon.
+@ruscio2004checklist describe three cases.
+First, when clinical and nonclinical participants are separately recruited and merged into a single sample.
+Second, when a sample is artificially split into subsamples based on different selection criteria, and these are then used to test different taxometric questions [cf.
+"subtractive" compound sampling, @haslam2020metaanalysis].
+Third, when researchers trim observations from the putative complement (individuals without a diagnosis) to inflate the base rate of the taxon.
 Each of these procedures risks imposing a categorical structure onto data that might be dimensional in nature.
-It is worth noting that artificial admixture poses a threat not only to taxometric analysis, but also to clustering methods, for the same reasons: both infer latent structure based on observed covariation, which can be distorted by selection artifacts.
+Importantly, admixture poses a threat not only to taxometric analysis, but also to clustering methods, for the exact same reason.
 
@@
 To address this, researchers draw a second sample consisting of an additional $n = 1000$, targeting clinically diagnosed cases, and merge it with the original sample, thereby creating an artificial admixture.
-The full simulation code and data are available in the online materials:
+The full simulation code and data are available in the online materials:\
 <https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba>
@@
 However, the issue appeared to be nuanced and less severe in this subset of studies.
-For example, @obrien2012dyslexia used psychometric cutoffs across different tools and inclusion criteria, which may not necessarily have biased the results unless the selection criteria correlate with different underlying cognitive impairments (phonological vs.
+For example, @obrien2012dyslexia used psychometric cutoffs across different tools and alternate inclusion criteria, which may not necessarily have biased the results, unless the selection criteria correlate with different underlying cognitive impairments in dyslexia (phonological vs.
 non-phonological).
@@
 Overall, the presence of artificial admixture was most frequently associated with taxonic conclusions.
-Only two notable mismatches were identified: @marcus2012latent reached a dimensional conclusion regarding ADHD despite having incurred admixture of clinical cases, while @munson2008asd reached a taxonic conclusion about ASD subtypes without clear evidence of admixture (all children had received an ASD diagnosis based on what appeared to be a uniform clinical assessment).
-As discussed above, a third exception was observed in a secondary analysis of a subsample by @frazier2010asd targeting ASD.
-Taken together, these findings reinforce the view that ADHD is best conceptualized as a dimensional condition, while ASD may exhibit taxonic features, although the latter conclusion is supported by limited unbiased evidence.
+Only two notable mismatches were identified: @marcus2012latent reached a dimensional conclusion regarding ADHD despite having incurred admixture of clinical cases, while @munson2008asd reached a taxonic conclusion about ASD subtypes without evidence of admixture (all children had received an ASD diagnosis based on a seemingly uniform clinical assessment).
+Together with the above discussed exception in a secondary analysis of a subsample by @frazier2010asd targeting ASD, these findings reinforce the view that ADHD is best conceptualized as dimensional, while ASD may exhibit taxonic features.
+However, the latter conclusion is supported by limited unbiased evidence.
 
 In our simulated example, we found that artificial admixture represents a major threat to the validity of taxometric analysis, and to clustering methods as well.
-One might argue that our simulated scenario, in which half of the diagnosed cases were separately sampled, was extreme.
+One might argue that our simulated scenario, in which half of sample was separately recruited, was extreme.
 However, this design is broadly consistent with the degree of admixture observed in the reviewed literature.
 It is worth noting that admixture is not limited to the obvious case of oversampling diagnosed individuals, and sometimes occurs in subtle ways.
-For example, through the undersampling of individuals with low trait levels, or by selectively drawing subsamples based on alternative clinical criteria within a larger dataset, as is often the case in subtype studies [cf.
-"subtractive" compound sampling, @haslam2020metaanalysis].
+For example, by selectively drawing subsamples based on alternative clinical criteria within a larger dataset, as is often the case in subtype studies [see "subtractive" compound sampling, @haslam2020metaanalysis].
 
@@
 While the DSM-5 may seem to endorse a categorical model, it remains atheoretical and does not explicitly or scientifically commit to a taxonic view.
-Researchers, rather than diagnostic manuals, may risk reifying these categories as natural taxa, thus overinterpreting their conceptual status, hence we advocate for more high-quality taxometric studies.
+Researchers, rather than diagnostic manuals, may risk reifying these categories as natural taxa, overinterpreting their conceptual status.
+Hence we advocate for more high-quality taxometric studies.
 
@@
 
-A first and obvious limitation of the present review concerns the small number of studies identified.
-This prevented us from conducting a meta-analysis: the available evidence was too scattered, methodologically heterogeneous, and inappropriate for quantitative synthesis.
-However, this limitation is itself informative, underscoring the need for more taxometric investigation in the field of NDDs.
-While the search query was comprehensive and no time limits were applied, more studies might have been included by relaxing the eligibility criteria.
+A first and obvious limitation of the present review concerns the small number of studies identified, and the virtual lack of studies targeting learning disorders as taxa.
+This prevented us from conducting a meta-analysis: the available evidence was too scattered and methodologically heterogeneous.
+However, this limitation is itself informative, underscoring the need for more taxometric investigation in the field.
+More studies might have been included by relaxing the eligibility criteria.
 However, this is unlikely to have made a substantial difference.
@@
 A relevant future direction for this line of investigation concerns a systematic examination of the use and results of cluster analysis.
-As described above, some of the reviewed articles employed clustering methods, and clustering is a prominent tool in this field [@astle2022trandiagnostic].
-We found that model-based approaches, such as latent class/profile analysis, were the most frequently used.
+Indeed, some of the reviewed articles employed clustering methods, and clustering is a prominent tool in this field [@astle2022trandiagnostic].
+Model-based approaches, such as latent class/profile analysis, were the most frequently used.
 The use of clustering is not without problems.
@@
 Second, meaningful scientific inference from clustering (beyond its simple use as a data reduction technique) requires careful handling of assumptions, as the risk of detecting pseudoclusters is high [@toffalini2022entia; @toffalini2024clusters].
-Third, it is is not immune to risks such as artificial admixture.
-Nonetheless, clustering and taxometric analysis could be more directly, but also critically, examined and compared when addressing the question of whether neurodevelopmental conditions reflect categories or dimensions.
+Finally, it is is not immune to risks such as artificial admixture.
+Nonetheless, clustering and taxometric analysis could be more directly, but also critically, compared when addressing the question of whether neurodevelopmental conditions reflect categories or dimensions.
 

### 2025-07-25 R/1 analysis-review-data.R

@@ -47,17 +47,26 @@ median(df$Sample_size)
 # median sample size by article (averaged)
 median((df |> group_by(ID_article) |> summarize(avg = mean(Sample_size)))$avg)
 
-# median number of indicators per analysis
+# number of indicators per analysis
+min(df$Indicators_number, na.rm = T)
+max(df$Indicators_number, na.rm = T)
 median(df$Indicators_number,na.rm=T)
 
 # methods used
 grepl("MAMBAC",df$Taxometric_methods_used)
 sc = str_count(df$Taxometric_methods_used,";")+1; min(sc); median(sc); max(sc); mean(sc>1)
 
+# taxonic conclusions
+sum(df$Taxonic_conclusion)
+sum(grepl("ASD", df$Target_disorder) & df$Taxonic_conclusion == 1) # total, concerning ASD
+round(100 * sum(grepl("ASD", df$Target_disorder) & df$Taxonic_conclusion == 1) / sum(grepl("ASD", df$Target_disorder)), 1) # average, concerning ASD
+
 # artificial admixture matching taxonic conclusions
 sum(df$Artificial_admixture)
-sum(df$Taxonic_conclusion)
-sum(df$Taxonic_conclusion[df$Artificial_admixture])
+round(100 * mean(df$Taxonic_conclusion[df$Artificial_admixture == 1]), 1) # percentage of studies reaching a taxonic conclusion, with artificial admixture
+round(100 * mean(df$Taxonic_conclusion[df$Artificial_admixture == 0]), 1) # percentage of studies reaching a taxonic conclusion, without artificial admixture
+
+# 
 
 ################################################################
 

### 2025-08-09 Paper/Paper.qmd
computationally reproducible paper redone

@@
     affiliation: University Vita-Salute San Raffaele, Milan, Italy
+  - name: Edmund Sonuga-Barke
+    corresponding: false
+    email: edmund.sonuga-barke@kcl.ac.uk
+    affiliation: King’s College London, UK
 author-note:
   disclosures:
-    study-registration: 'The review was preregistered at https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0'
+    study-registration: 'Review preregistered protocol: https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0'
     financial-support: 'Financial support: Project 2022KBW99S "Tails or types?
@@
     authorship-agreements: null
-abstract: "Taxometric analysis is explicitly designed to test whether the latent structures underlying clinical psychological constructs are categorical (taxonic) or dimensional.
-Despite its relevance to ongoing debates about the nature of neurodevelopmental disorders (NDDs), its application in this field has been limited.
-We conducted a preregistered systematic review to critically evaluate the literature, with a focus on methodological threats such as artificial admixture.
-All published taxometric studies on autism, ADHD, learning disorders, and language impairment were identified and assessed.
-Studies on ADHD and language impairment generally supported dimensional structures, whereas most studies on autism and NDD subtypes predominantly reported taxonic findings.
-However, artificial admixture was frequently observed and undermines the credibility of many taxonic conclusions.
-To illustrate this risk, we conducted a Monte Carlo simulation showing how admixture can spuriously generate pseudotaxonic results.
-We urge caution in interpreting the outcomes, and advocate for more high-quality taxometric studies in this area."
+abstract: "**Background.** Taxometric analysis is specifically designed to test whether the latent structures underlying clinical conditions are categorical (taxonic) or dimensional in nature.
+Despite its relevance to ongoing debates about the nature of neurodevelopmental conditions, its application to this field has been limited.
+**Methods.** We conducted a preregistered systematic review of published taxometric studies on autism, ADHD, learning disorders, and language impairment, with a focus on methodological threats such as artificial admixture (i.e., the practice of combining subsamples recruited via alternative methods, such as clinical and control groups, into a single dataset).
+A Monte Carlo simulation was run to illustrate how admixture can yield false taxonic results from dimensional data.
+**Results.** Fourteen articles met the inclusion criteria.
+Studies of ADHD and language impairment largely supported dimensional models, whereas most studies of autism and conditions’ subtypes supported categorical models.
+However, admixture was common in studies reaching taxonic conclusions, especially for autism.
+In simulations, admixture was consistently associated with taxonic findings despite the true structure being dimensional.
+**Conclusions.** Evidence for categorical structure, especially in autism and disorders’ subtypes, is likely undermined by methodological limitations, particularly admixture.
+We advocate for more high-quality taxometric studies in this area."
 keywords: [Neurodevelopmental disorders, Taxometric analysis, Autism, ADHD, Learning disorders]
@@
 
-In recent years, research on neurodevelopmental disorders (NDDs) has undergone significant reconsideration.
-Traditional diagnostic categories defined in the DSM-5 [@dsm5] have increasingly been questioned in favor of a dimensional and transdiagnostic perspective [e.g., @astle2022trandiagnostic; @michelini2024world; @happe2020looking; @happe2021both; @sonuga2020kuhnian; @caviola2024editorial].
-Notably, the DSM-5 itself has incorporated some dimensional features (for example by defining a broad Autism Spectrum Disorder \[ASD\] in place of distinct diagnoses).
-Several factors have motivated this shift: the marked heterogeneity within diagnostic groups, the weak correspondence between data-driven clusters and DSM-based categories, the frequent comorbidity and phenotypic correlations across conditions, and the view that individuals with NDDs may represent the tails of developmental continua, quantitatively rather than qualitatively different from the so-called neurotypical population [@michelini2024world; @happe2021both; @mammarella2021nocore].
+In recent years, there has been a reconsideration of the conceptual foundations of neurodevelopmental conditions (NDCs).
+The notion that diagnostic entities defined in the DSM-5 reflect discrete categories with identifiable non-arbitrary ‘natural’ boundaries,   separating disorder from non-disorder, and one disorder from another [@dsm5] has increasingly been questioned in favor of dimensional and transdiagnostic perspectives [e.g., @astle2022trandiagnostic; @caviola2024editorial; @happe2020looking; @happe2021both; @mammarella2021nocore; @michelini2024world; @posner2020adhd; @sonuga2020kuhnian].
+In fact, the issue has been discussed for a long time.
+@sonuga-barke1998categoricalmodels questioned whether a medical model that frames childhood conditions as categorical disorders intrinsically linked to dysfunction was appropriate and useful.
+While diagnostic manuals such as the DSM are presented as atheoretical, they in practice adopt a "neo-Kraepelinian" vision of psychopathology, treating disorders as natural disease entities [see also @coghill2012categories].
 
-Determining whether NDDs represent categories, rather than just tails of continuous distributions spanning the general population, is important for several reasons, including questions of social and personal identity for affected individuals, implications for diagnostic framing, comorbidity, and intervention choices in clinical contexts, and consequences for the organization of classrooms, educational practices, and even the design and power analysis of research studies [@caviola2024editorial].
-More fundamentally, this question is relevant for our understanding of the ontological status of psychological phenomena.
+A dimensional perspective reinterprets NDCs as extreme variations along continuous multivariate traits that are at least quasi-normally distributed in the general population.
+Accordingly, individuals with NDCs are not categorically distinct, and differ quantitatively rather than qualitatively from the neurotypical population.
+From this perspective diagnostic thresholds are essentially arbitrary, determined not by natural discontinuities but by pragmatic considerations such as clinical need [@arildskov2022howmuch; @haslam2006latentstructure].
 
-A dimensional approach aligns well with long-standing principles in the study of individual differences in psychology, and has long characterized research on intelligence and personality [@caviola2024editorial].
-It also seems well-suited for NDDs, given a growing body of evidence supporting polygenicity and multifactorial etiology [@plomin2005generalist; @kovas2006generalist; @pennington2012multiple; @astle2020core; @caviola2024editorial; @demontis2023adhdgenome].
-Most behavioral traits are understood to result not from one or a few causes, but from the additive effects of many small influences [@chabris2015fourth].
-According to the central limit theorem [@fischer2011central], when many individual causes (e.g., discrete genetic variants) contribute additively, the resulting trait tends to approximate a continuous, Gaussian distribution.
-Nonetheless, a principled categorical view still resists in neurodevelopmental research, for example concerning ASD [e.g., @happe2021both; @chown2021bit; @frazier2023asd], subtypes within NDDs [e.g., @burgess2018taxodyslexia] or the proposed emergence of data-driven taxa [@astle2022trandiagnostic].
+Determining whether NDCs represent tails of continuous distributions or discrete categories with boundaries grounded in biology or clinical reality, is important for our understanding of the ontological status of neurodevelopmental phenomena and for assessing the adequacy of the medical dysfunction paradigm.
+If NDCs are dimensional in nature, then the diagnostic thresholds used to categorize disorders are, by definition, arbitrary, reframing diagnostic categories as socially constructed groupings that include individuals at the margins of trait continua.
+The implications potentially affect how diagnosed individuals understand their social identity, how clinicians make decisions, how classrooms and support systems are organized to accommodate neurodivergence, and how research is designed and interpreted [@caviola2024editorial].
 
-Taxometric analysis is a set of quantitative methods originally introduced by Paul Meehl and colleagues and refined over decades [@meehl1995taxometric], explicitly aimed at establishing whether a clinical disorder is characterized by an underlying categorical or dimensional structure.
-It has been widely applied in psychopathology, and typically suggested that most mental health conditions (e.g., depression, anxiety) are better conceptualized as dimensional [@haslam2012review; @haslam2020metaanalysis].
-In brief, taxometric procedures, such as MAMBAC, MAXCOV, MAXEIG, and L-Mode examine covariation patterns among observed indicators to adjudicate between competing latent categorical vs dimensional models [@ruscio2007taxometric; @ruscio2004checklist; @cole2004taxometric].
-These methods now feature simulation-based tools like the Comparison Curve Fit Index (CCFI), which provide quantitative measures to facilitate inference [@ruscio2018using].
+Support for a dimensional account of NDCs arises from several converging lines of evidence.
+First, individuals sharing the same diagnostic label often show substantial heterogeneity in symptom profiles, cognitive functioning, and developmental trajectories [e.g., @astle2022trandiagnostic].
+Second, the boundaries between diagnostic categories are frequently blurred, with high comorbidity [@coghill2012categories; @michelini2024world].
+Third, dimensional models are consistent with the growing evidence supporting multifactorial and polygenic etiology in NDCs [@astle2020core; @caviola2024editorial; @kovas2006generalist; @pennington2012multiple; @plomin2005generalist].
+According to the central limit theorem [@fischer2011central], when many individual causes (e.g., genetic variants) contribute additively, the resulting trait tends to approximate a continuous, Gaussian distribution.
+Despite this body of evidence, categorical thinking still dominates parts of the neurodevelopmental literature, for example concerning the categorical nature of ASD [e.g., @chown2021bit; @frazier2023asd; @happe2021both], subtypes within NDCs [e.g., @burgess2018taxodyslexia] or the proposed interpretation of emerging data-driven taxa [@astle2022trandiagnostic].
 
-Unfortunately, the use of taxometric analysis has remained limited in neurodevelopmental research, with most existing studies focusing on adult psychopathology [@beauchaine2003taxometrics; @haslam2020metaanalysis].
-In their debate on whether ASD is best conceptualized as a category or a continuum, @happe2021both and @chown2021bit do not even mention such methods.
-In their influential review, @astle2022trandiagnostic briefly acknowledge taxometry, but predominantly emphasize studies using cluster analysis.
+While much of the evidence for dimensionality is circumstantial, taxometric analyses, originally introduced by Paul Meehl and colleagues [@meehl1995taxometric], were developed to provide a direct empirical test of whether latent clinical construct are categorical (taxonic) or dimensional in nature.
+Taxometric methods evaluate the structure of covariance among observed variables to determine whether there exists a latent boundary that divides individuals into qualitatively distinct groups.
+Conceptually, if such a boundary exists, it should alter and disrupt the patterns of relationships between indicators.
+Otherwise, the patterns are smooth and continuous across the entire range of observed values, supporting a dimensional model.
+Frequently used taxometric procedures include MAMBAC (Mean Above Minus Below A Cut), MAXCOV (Maximum Covariance), MAXEIG (Maximum Eigenvalue), and L-Mode (Latent Mode Factor Analysis) [@cole2004taxometric; @ruscio2007taxometric; @ruscio2004boundaryissues].
+Recently, simulation-based tools such as the Comparison Curve Fit Index (CCFI) have been introduced to provide objective, quantitative indices of fit between observed data and simulated taxonic versus dimensional comparison curves, facilitating inference [@ruscio2018using].
+
+Taxometric analyses have been widely applied in adult psychopathology, generally supporting dimensional structures for most adult psychiatric conditions [@haslam2012review; @haslam2020metaanalysis].
+However, their use in neurodevelopmental research has remained limited, though the importance of taxometric methods as explicit tests for determining whether childhood disorders are categorical or dimensional has been repeatedly emphasized [@sonuga-barke1998categoricalmodels; @beauchaine2003taxometrics; @coghill2012categories].
+In their debate on whether ASD reflects a category or a continuum, @happe2021both and @chown2021bit do not mention such methods.
+In their review, @astle2022trandiagnostic briefly acknowledge taxometry, but predominantly emphasize studies using cluster analysis (a family of data-driven methods aimed at grouping individuals based on similarity across multiple variables).
 However, the limitations of clustering methods for assessing taxonicity have been discussed for decades: clustering algorithms tend to impose (rather than test for) categorical structures, they provide weak tests of taxonicity, and often fail to recover known taxa in simulations [@beauchaine2003taxometrics; @ruscio2004boundaryissues].
-Recent simulations using psychologically realistic data (featuring both some skewness and intercorrelated indicators) suggest that commonly used clustering methods are prone to detecting pseudoclusters while presenting low power to identify true taxa [@toffalini2022entia; @toffalini2024clusters].
-Indeed, @astle2022trandiagnostic interpret clustering results as evidence that data-driven groups do not align with DSM-5 categories, rather than as compelling tests of latent structural form.
+Recent simulations using psychometrically realistic data suggest that commonly used clustering methods are prone to inflate the number of detected clusters [@toffalini2022entia; @toffalini2024clusters].
+Indeed, @astle2022trandiagnostic interpreted clustering results as evidence that data-driven groups fail to align with DSM-5 categories, rather than as strong tests of taxonicity.
 
-Taxometric analysis is not without limitations, some of which are shared with cluster analysis.
-Commonly cited threats to validity of conclusions include poor indicator validity, nuisance covariance, low base rate, and artificial admixture [@cole2004taxometric; @haslam2020metaanalysis; @ruscio2004checklist; @ruscio2018using].
-
-Poor indicator validity implies that indicators present little mean difference between individuals with and without the target condition (i.e., the putative taxon and complement).
-Ideally, Cohen’s *d* for each indicator should substantially exceed 1 (*d* $\gg$ 1).
-Nuisance covariance refers to strong residual correlations among indicators after accounting for group membership, indicating redundancy.
-Low base rate means that too few individuals in the sample belong to the taxon, limiting power (samples with *n* \< 30 for the taxonic group are usually deemed insufficient).
-This is particularly relevant when studying rare conditions.
+Like any method, taxometric analysis has its limitations.
+Commonly cited sources of bias in taxometric conclusions include poor indicator validity, nuisance covariance, low base rate, and artificial admixture [@cole2004taxometric; @haslam2020metaanalysis; @ruscio2018using; @ruscio2004boundaryissues; @ruscio2004checklist].
+Poor indicator validity implies that indicators present little mean difference between individuals with and without the target condition.
+Ideally, Cohen’s *d* for each indicator should substantially exceed 1.
+Nuisance covariance refers to strong residual correlations among indicators after accounting for group membership, indicating redundant information.
+Low base rate means that too few individuals in the sample belong to the taxon, limiting power to detect it (*n \>* 30 for the taxonic group is usually considered necessary), which is problematic when studying rare conditions.
 Artificial admixture occurs when participants are combined from different recruitment sources or selection criteria, often as an attempt to compensate for a low base rate.
-For instance, clinical cases may be recruited separately from controls and merged into a single sample.
-In general, poor indicator validity and low base rate reduce power to detect true taxa, while high nuisance covariance and artificial admixture increase the risk of detecting pseudotaxa by mimicking a categorical structure.
+@ruscio2004checklist describe three admixture-related sampling practices that may lead to pseudotaxonic results.
+First, when clinical and nonclinical participants are separately recruited and merged into a single sample.
+Second, when a sample is artificially split into subsamples based on different selection criteria, and these are then used to test different taxometric questions [cf.
+"subtractive" compound sampling, @haslam2020metaanalysis].
+Third, when researchers trim observations from the putative complement (individuals without a diagnosis) to inflate the base rate of the taxon.
+All these scenarios tend to artificially impose categorical structures onto data that might be strictly dimensional.
 
@@
 
-The primary aim of this systematic review is to critically examine the use of taxometric methods in studies of NDDs within the published literature.
-We focused on the most commonly investigated NDD categories: Autism Spectrum Disorder (ASD), Attention-Deficit/Hyperactivity Disorder (ADHD), Specific Learning Disorders (SLDs), and the closely related condition of Language Disorder/Impairment.
-(The latter was not included in the preregistered protocol and constitutes a minor deviation.) We excluded Intellectual Disability, as its etiology is often clearly identifiable and taxonic/syndromic by definition in many cases (e.g., Down syndrome), or linked to specific medical risk factors, while residual non-syndromic cases are typically regarded as reflecting the lower tail of general mental ability and thus inherently dimensional.
+We aimed to critically review the use of taxometric methods in published studies of NDCs.
+We focused on the most commonly investigated categories: Autism Spectrum Disorder (ASD), Attention-Deficit/Hyperactivity Disorder (ADHD), Learning Disorders, and the closely related condition of Language Impairment/Disorder (the latter constitutes a minor deviation from the preregistration protocol).
+We excluded Intellectual Disability, whose etiology is often identifiable and taxonic/syndromic by definition (e.g., Down syndrome), or linked to known medical risk factors.
 
 A crucial goal was to assess the methodological quality and credibility of taxometric findings.
-We specifically evaluated potential sources of bias, such as artificial admixture, and assessed whether studies provided sufficient methodological information to evaluate other known threats to validity, including low indicator validity, high nuisance covariance, and skewed indicator distributions.
-(Strong skewness can create the illusion of a categorical structure, and it might be relevant when assessing rare or extreme behaviors.)
+We specifically evaluated potential sources of bias, such as artificial admixture, and assessed whether studies provided sufficient methodological information to evaluate other known threats to validity, including low indicator validity, high nuisance covariance, and indicators skewness [extreme skewness can also create the illusion of categorical structures, @ruscio2004boundaryissues; @ruscio2004checklist, and it might be relevant when assessing rare or extreme behaviors].
 
-Following the review, we report a Monte Carlo simulation study designed to empirically illustrate the extent to which artificial admixture can lead to the spurious detection of categorical latent structures in otherwise dimensional data.
+Finally, we conduct a Monte Carlo simulation study to empirically illustrate the extent to which artificial admixture can lead to the risk to reaching taxonic conclusions in otherwise dimensional data.
+We simulate a dimensional scenario where clinical and nonclinical participants are separately recruited and merged into a single sample; taxometric (and model-based cluster) analyses are then performed on a set of observed indicators reflecting the latent clinical dimension.
 
-# Systematic Review of Taxometric Analysis on NDDs
+# Methods
 
-## Methods
-
-The systematic review was conducted in accordance with the Preferred Reporting items for Systematic Reviews and Meta-Analysis (PRISMA) 2020 guidelines.
-A review protocol was preregistered on OSF registries at the following link: https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0 The methods are detailed below.
+The systematic review was conducted in accordance with the *Preferred Reporting items for Systematic Reviews and Meta-Analysis* (PRISMA) 2020 guidelines.
+A review protocol was preregistered on OSF registries at the following link: https://osf.io/65y9g/?view_only=91e97c6b00dc444abb83fefcf67cbac0
 
@@
 A systematic search for relevant peer-reviewed articles was conducted in April 2025 using four electronic databases: PsycINFO, PubMed, Scopus, and Web of Science.
-No time limits were applied, although records predating 1990 were not expected, as taxometric methods were not yet widely adopted at that time.
-The following search query—reflecting the methods and target populations of interest—was applied to titles, abstracts, and keywords:
+No time limits were applied.
+The following search query, reflecting the methods and populations of interest, was applied to titles, abstracts, keywords:
 
@@
 
-In databases where the wildcard `*` was not supported (e.g., PubMed), it was replaced with full terms (e.g., `"disorder"`, `"disability"` instead of `"dis*"`; `"mathematical"`, `"mathematics"`, `"math"` instead of `"math*"`).
+In PubMed, where the wildcard `*` was not supported, it was replaced with full terms (e.g., `"disorder"`, `"disability"` instead of `"dis*"`; `"mathematical"`, `"mathematics"`, `"math"` instead of `"math*"`).
 
@@
 References cited in, and articles citing, @haslam2020metaanalysis were also screened via Google Scholar.
-Finally, an exploratory search using OpenAI’s GPT-4o was conducted, using the preregistration protocol as contextual input.
-All promising records identified through these additional methods and not retrieved via database searches were listed under the *"Identification of studies via other methods"* section of the PRISMA flowchart.
+Finally, an exploratory search using OpenAI’s GPT-4o “Deep Research” tool was conducted, providing the preregistration protocol as context.
+All additional promising records identified through these methods were listed under the *“Identification of studies via other methods”* section of the flowchart.
 
@@
 
-Inclusion criteria were as follows:\
-1. The study applied at least one taxometric procedure;\
-2. The target population consisted of individuals with NDDs or related traits;\
-3. The study drew a conclusion about the latent structure (categorical vs. dimensional), or reported sufficient information (e.g., CCFIs) to infer such a conclusion.
+Inclusion criteria were:
 
-Exclusion criteria were:\
-1. The article was theoretical, methodological, or a simulation study without empirical data;\
-2. The study did not report any behavioral data (in practice, no screened study was excluded for this criterion);\
-3. The study was a review and/or meta-analysis;\
-4. The study employed latent structure methods other than taxometric analysis (e.g., only clustering), with no taxometric procedures reported.
+1.  The study applied at least one taxometric procedure (MAMBAC, MAXCOV, MAXEIG, L-Mode, MAXSLOPE; other procedures would be considered if explicitly described as taxometric);
+2.  The target population consisted of individuals with NDCs/traits;
+3.  The study drew a conclusion about the latent structure (categorical vs. dimensional) or reported sufficient information (e.g., CCFIs) to infer a conclusion.
 
-In addition, following the preregistered protocol, the search was limited to published, peer-reviewed journal articles.
-This decision reflected the aim of providing a critical review of the existing empirical literature.
-Only one potentially relevant non-journal record was identified and, while excluded from the systematic review, is discussed separately in the *Limitation* section.
+Exclusion criteria were:
+
+1.  The article was theoretical, methodological, simulation study without empirical data;
+2.  The study did not analyze behavioral data;
+3.  The study was a review and/or meta-analysis.
+
+The search was limited to published, peer-reviewed journal articles, as we aimed to provide a critical review of the existing high-quality literature.
+Only one potentially relevant non-journal record was identified and, while excluded from the systematic review, is commented separately in the Discussion.
 
@@
 
-Two independent human coders (MC and RP) screened and coded all articles.
-No AI tools conducted any screening or coding, although coders occasionally used them to help clarify specific article content.
-All discrepancies were resolved through discussion and were later reviewed by the PI (ET).
+Two independent human coders (MC and RP) screened and coded all articles. All discrepancies were resolved through discussion and were later reviewed by the PI (ET).
 
-Since a single article could report multiple (sub)samples and/or multiple taxometric analyses on the same sample, the full coded dataset followed a long-form data structure, with some articles appearing across multiple rows.
+Since a single article could report multiple (sub)samples and/or taxometric analyses, the full coded dataset followed a long-form data structure, with articles appearing across multiple rows.
 Coded variables were defined in the data dictionary included in the preregistration protocol, with a few additional fields added to capture coder notes and agreement metrics.
@@
 Inter-coder agreement was fair to good across all phases.
-In the title and abstract screening phase, Cohen’s *κ* = 0.79 (34 out of 38 independent decisions matched).
-In the full-text eligibility screening, Cohen’s *κ* = 0.52 (15 out of 19).
+In the title and abstract screening phase, Cohen’s *κ*=0.79 (34 out of 38 independent decisions matched).
+In the full-text eligibility screening, Cohen’s *κ*=0.52 (15 out of 19).
 For the final data extraction, 84% of rows (71 out of 85) were extracted by both coders.
-Agreement on critical variables was high: for the "authors' taxonic conclusion" agreement was 94%; for "artificial admixture", it was 95% at the sample or subsample level.
+Agreement on critical variables was high: for the “authors’ taxonic conclusion” agreement was 94%; for “artificial admixture”, it was 95% at the (sub)sample level.
 
-## Results
+# Results
 
 A total of 110 records were identified through database searches.
-After removing 67 duplicates and excluding 5 additional records (dissertations or book chapters, only one of which appeared to include primary taxometric analyses but could not be retrieved), 38 unique articles were screened at the title and abstract level.
-
+After removing 67 duplicates and excluding 5 additional records (dissertations or book chapters, only one of which appeared to include primary taxometric analyses but could not be retrieved), 38 articles were screened for title/abstract.
 Nineteen records were excluded at this stage.
-Full-text assessment was then conducted on the remaining 19 articles, of which 7 were excluded for not meeting the inclusion criteria: 4 did not apply taxometric methods, and 3 did not target neurodevelopmental disorders.
-
-An additional 4 records were identified through citation tracking and online browsing. Of these, 2 were excluded (1 was a methodological article, and 1 did not employ taxometric analysis).
-
-In total, 14 studies met inclusion criteria and were included in the review. The PRISMA flowchart in @fig-prisma summarizes the study selection process.
+Full-text assessment was then conducted on the remaining 19 articles, of which 7 were excluded for not meeting the inclusion criteria.
+An additional 4 records were identified through citation tracking and online browsing, of which 2 were included in the final review.
+The PRISMA flowchart in @fig-prisma summarizes the study selection process.
 
@@
 A full reporting of all coded data for each individual taxometric analysis is available in the Supplemental Online Material.
-The total number of taxometric analyses was `r nrow(df)`, ranging from `r min(table(df$ID_article))` to `r max(table(df$ID_article))` analyses per article.
-The targeted disorders were ADHD (6 articles), ASD (6 articles), dyslexia (1 article), and Specific Language Impairment (SLI; 2 articles), with one article addressing both ADHD and ASD.
-Most articles (10 out of 14) examined whether a given disorder constituted a taxon, while the remaining 4 focused on whether putative subtypes within a disorder exhibited taxonic structure.
+The total number of taxometric analyses was `r nrow(df)`, ranging from `r min(table(df$ID_article))` to `r max(table(df$ID_article))` per article.
+Targeted disorders were ADHD (6 articles), ASD (6 articles), dyslexia (1 article), and Specific Language Impairment (SLI; 2 articles), with one article addressing both ADHD and ASD.
+Most articles (10 out of 14) examined whether a given disorder constituted a taxon, while the remaining 4 focused on whether putative subtypes within a disorder showed taxonic structure.
 
@@
 In contrast, all four articles examining ASD as a disorder leaned toward, or explicitly supported, a categorical (taxonic) structure.
-The remaining four articles investigated subtypes of disorders, and all leaned toward taxonic conclusions—either strongly or predominantly [@stevens2018functional on ADHD subtypes; @munson2008asd on ASD subtypes; @obrien2012dyslexia on dyslexia subtypes], or partially [@ingram2008defining, which supported ASD subtypes as taxa depending on the set of indicators used].
+The remaining four articles investigated subtypes of disorders, and all favored taxonic conclusions, either predominantly [@stevens2018functional on ADHD subtypes; @munson2008asd on ASD subtypes; @obrien2012dyslexia on dyslexia subtypes], or partially [@ingram2008defining supported taxonic ASD subtypes depending on the set of indicators used].
 In total, 8 out of 14 articles reached a taxonic conclusion in at least some of their taxometric analyses.
@@
 
-Artificial admixture was evident or strongly suspected in at least part of the analyses in 8 out of 14 articles, involving a total of `r sum(df$Artificial_admixture)` out of `r nrow(df)` analyses.
-Notably, articles that incurred artificial admixture tended to align with those that reached taxonic conclusions: 6 of the 8 articles with suspected admixture also reported taxonic findings.
+Artificial admixture was evident or strongly suspected in at least part of the analyses in 8 out of 14 articles (including all those targeting ASD) and involving a total of `r sum(df$Artificial_admixture)` out of `r nrow(df)` analyses.
+Articles that incurred artificial admixture tended to align with those that reached taxonic conclusions: 6 of the 8 articles with suspected admixture also reported taxonic findings.
 This association was even more pronounced at the level of individual analyses.
 A taxonic conclusion was reached in `r sum(df$Taxonic_conclusion[df$Artificial_admixture == 1])` out of `r sum(df$Artificial_admixture == 1)` analyses (`r round(100 * mean(df$Taxonic_conclusion[df$Artificial_admixture == 1]), 1)`%) where admixture was evident or suspected.
-In contrast, only `r sum(df$Taxonic_conclusion[df$Artificial_admixture == 0])` out of `r sum(df$Artificial_admixture == 0)` analyses (`r round(100 * mean(df$Taxonic_conclusion[df$Artificial_admixture == 0]), 1)`%) reached a taxonic conclusion when admixture was likely or certainly absent.
+In contrast, only `r sum(df$Taxonic_conclusion[df$Artificial_admixture == 0])` out of `r sum(df$Artificial_admixture == 0)` analyses (`r round(100 * mean(df$Taxonic_conclusion[df$Artificial_admixture == 0]), 1)`%) reached a taxonic conclusion when admixture was likely or certainly
 
-Remaining methodological issues concerned missing information about key descriptive statistics and validity indices.
-In particular, **nuisance covariance**—that is, correlations among indicators within diagnostic groups or within the putative taxon and complement—could be coded for only 2 out of 14 articles.
-Where reported, it was small to negligible in @james2016latent, but exceeded 0.40 for many indicators in @frazier2023asd; notably, both studies reached a taxonic conclusion.
-**Skewness** of indicator score distributions was reported in only 5 articles.
-When available, skewness values were often near or beyond ±1 for at least one indicator in roughly one-third of analyses.
+Other methodological issues concerned missing information about key descriptive statistics and validity indices.
+Nuisance covariance could be coded for only 2 out of 14 articles.
+Where reported, it was small to negligible in @james2016latent , but exceeded 0.40 for many indicators in @frazier2023asd .
+Both studies reached a taxonic conclusion.
+Skewness of indicator score distributions was reported in only 5 articles, where skewness values were often near or beyond ±1 for at least one indicator in about one-third of analyses.
 However, there was no clear association between skewness and whether a taxonic conclusion was reached.
-By contrast, **indicator validity** was more consistently reported, and could be coded in all but 3 articles.
-Where available, validity (typically expressed as Cohen’s *d*) was almost always large or very large (*d* $\gg$ 1), consistent with best-practice recommendations for taxometric analysis.
+By contrast, indicator validity was more consistently reported, and could be coded in all but 3 articles.
+Where available, it was almost always large or very large (Cohen’s *d* ≫ 1), consistent with best-practice recommendations for taxometric analysis.
 
@@
 
-Among the 14 identified articles, 6 also employed clustering methods—typically latent class analysis (LCA) or latent profile analysis (LPA).
-In 5 of these 6 cases, clustering was conducted in parallel with taxometric analysis [@james2016latent; @munson2008asd; @frazier2023asd; @frazier2007adhd; @deserno2023probing], while in one case it was presented as a secondary analysis [@frazier2010asd].
+Among the 14 identified articles, 6 also employed clustering methods: in 5 cases consisting of latent class/profile analysis (LCA, LPA); the remaining case employing community detection in a graph analysis.
+In 5 of these 6 articles clustering was conducted in parallel with taxometric analysis [@deserno2023probing; @frazier2007adhd; @frazier2023asd; @james2016latent; @munson2008asd], while in one case it represented a secondary analysis [@frazier2010asd].
 All six studies targeted ADHD and/or ASD, and in all cases clustering results were interpreted as supporting taxonic conclusions.
-These interpretations generally converged with those from taxometric analysis, with one exception: in @frazier2007adhd, clustering results were partially inconclusive but suggested the presence of subgroups within ADHD, thus contrasting with the dimensional structure favored by the taxometric analysis.
+These interpretations generally converged with those from taxometric analysis, with one exception: in @frazier2007adhd, clustering results inconclusively suggested the presence of subgroups within ADHD, contrasting with the dimensional structure favored by the taxometric analysis (it is worth reminding that clustering methods do not provide explicit tests of taxonicity).
 In all but two studies [@frazier2007adhd; @munson2008asd], we identified at least some risk of artificial admixture.
@@
 
-Artificial admixture represents a powerful threat to validity that is common but also easily avoidable by design.
-Different types of admixture can be incurred.
-@ruscio2004checklist describe three cases.
-First, when clinical and nonclinical participants are separately recruited and merged into a single sample.
-Second, when a sample is artificially split into subsamples based on different selection criteria, and these are then used to test different taxometric questions [cf.
-"subtractive" compound sampling, @haslam2020metaanalysis].
-Third, when researchers trim observations from the putative complement (individuals without a diagnosis) to inflate the base rate of the taxon.
-Each of these procedures risks imposing a categorical structure onto data that might be dimensional in nature.
-Importantly, admixture poses a threat not only to taxometric analysis, but also to clustering methods, for the exact same reason.
+Artificial admixture represents a common and powerful threat to validity, but also one that is easily avoidable by design.
+In our review, it presented a marked risk of bias for studies supporting the taxonic nature of autism.
+To illustrate the risks associated with artificial admixture, we simulate a dimensional scenario where clinical and nonclinical participants are separately recruited and merged into a single sample
 
@@
 
-We simulate a dimensional scenario to visually demonstrate how artificial admixture, specifically of the first type described above, can produce pseudotaxonic results.
-A single normally distributed latent variable $X$ is generated to represent the underlying clinical trait of interest.
-A set of observed indicators ($x_1$, $x_2$, $x3$) are then simulated as linear functions of $X$ plus normally distributed random noise, yielding an $R^2$ of 0.77 for each indicator.
-Individuals above the 99th percentile of $X$ are assumed to receive a diagnosis via global clinical assessment.
+A single normally distributed latent variable $X$ is generated to represent the clinical trait of interest.
+A set of observed indicators ($x_1$, $x_2$, $x_3$) are then simulated as linear functions of $X$ plus normally distributed random noise, yielding $R^2$ = 0.77 for each indicator.
+Individuals above the 99^th^ percentile of $X$ are assumed to receive a diagnosis via global clinical assessment.
 A sample of $N = 1000$ individuals is drawn from the general population.
 Given the low base rate, only about $n = 10$ cases are expected to naturally receive a diagnosis.
-To address this, researchers draw a second sample consisting of an additional $n = 1000$, targeting clinically diagnosed cases, and merge it with the original sample, thereby creating an artificial admixture.
+To address this, researchers draw a second sample consisting of an additional $n = 1000$, targeting clinically diagnosed cases, and merge it with the original sample.
 The full simulation code and data are available in the online materials:\
-<https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba>
+[https://osf.io/ys5ad/?view_only=0e2812acfabf43fc8af01f883c7331ba](#0){.uri}
 
@@
 However, nuisance covariance is partly problematic: among diagnosed individuals, the correlation between indicators is close to zero ($r \approx 0.05$), whereas among non-diagnosed individuals, it is substantial ($r \approx 0.60$).
-This asymmetry, combined with differential recruitment, contributes to the illusion of a latent taxon.
 
@@
 
-A Monte Carlo simulation was conducted with 1000 iterations to assess the impact of artificial admixture on taxometric outcomes.
-Analyses were performed using the `RTaxometric` package in R.
-As recommended in the literature, inference was based on the Comparison Curve Fit Index (CCFI), which ranges from 0 (indicating strong support for a dimensional structure) to 1 (indicating strong support for a categorical structure), with values around 0.5 considered ambiguous.
+A Monte Carlo simulation was then conducted with 1000 iterations.
+Taxometric analyses were performed using the `RTaxometric` package in R.
+Inference was based on the Comparison Curve Fit Index (CCFI), which ranges from 0 (indicating strong support for a dimensional structure) to 1 (strong support for a categorical structure), with values around 0.5 considered inconclusive.
 The default procedures (MAMBAC, MAXEIG, and L-Mode) were applied to each simulated dataset.
@@
 
-Thus, without admixture taxometric results leaned towards a dimensional conclusion, but frequently remained inconclusive.
-This is probably due to nuisance covariance.
-It is worth noting, however, that under a dimensional structure achieving both high indicator validity and low within-group covariance might be practically impossible.
-This is because indicators that clearly differentiate individuals with and without a diagnosis must be strongly associated with the underlying trait, an association that produces inter-indicator covariance.
-In fact, this trade-off may iteself serve as indirect evidence for dimensionality.
+Thus, without admixture taxometric results leaned towards a dimensional conclusion, but frequently remained inconclusive, probably due to nuisance covariance.
+It is worth noting, however, that under a dimensional structure achieving both high indicator validity and low within-group covariance might be practically impossible, because indicators that strongly differentiate individuals with and without a diagnosis must be strongly associated with the underlying trait, thus producing inter-indicator covariance.
 
@@
 
-We found that taxometric analysis remains underused in neurodevelopmental research, which is unfortunate given the unique relevance of this method to the ongoing debate about the nature of NDDs [@astle2022trandiagnostic; @michelini2024world; @happe2021both; @sonuga2020kuhnian].
+We found that taxometric analysis remains underused in neurodevelopmental research, which is unfortunate given the unique relevance of this method to the ongoing debate about the nature of NDCs [@astle2022trandiagnostic; @happe2021both; @michelini2024world; @sonuga-barke1998categoricalmodels; @sonuga2020kuhnian].
 That underuse was expected, however.
-@haslam2020metaanalysis classified only 15 out of 183 taxometric studies in the "childhood" category (which included heterogeneous conditions, and even sleep problems).
+@haslam2020metaanalysis classified only 15 out of 183 taxometric studies in the “childhood” category (which included heterogeneous conditions, and even sleep problems).
 In the studies identified in our review, only ASD, ADHD, and SLI were targeted as disorders in taxometric analysis.
-None of learning disorders was targeted as a taxon, with Dyslexia being covered in one single article, but only for investigating its subtypes.
+None of learning disorders employed taxometric approaches, with Dyslexia being covered in one single article, but only for investigating its subtypes.
 
@@
 Dimensional conclusions were predominantly reached for ADHD and SLI, while categorical conclusions prevailed for ASD and studies examining subtypes of any disorder.
-This echoes @haslam2020metaanalysis, and may seem to offer a compelling answer to the debate raised by @chown2021bit and @happe2021both.
-However, a closer inspection of the methodological limitations in the reviewed studies suggests that such conclusions should be interpreted with caution.
+This may seem to offer a compelling answer to the debate raised by @chown2021bit and @happe2021both However, a closer inspection of the methodological limitations in the reviewed studies suggests that such conclusions should be interpreted with caution.
 
-We identified four studies that targeted ASD as a disorder in taxometric analyses, and all of them reached taxonic conclusions.
-However, some artificial admixture was identified in all of them, and it represents a major threat to conclusion validity, as demonstrated both in our data simulation and in prior literature [@ruscio2004checklist; @ruscio2004boundaryissues].
-Admixture in the reviewed studies ranged from sampling participants through different methods, only some of which explicitly targeted clinical populations [@james2016latent], to combining clinical and non-clinical samples [@frazier2023asd; @deserno2023probing].
+We identified four ASD taxometric analyses, and all of them reached taxonic conclusions, with some degree of artificial admixture.
+This represents a major threat to validity of the conclusions of these studies, as demonstrated both in our data simulation and in prior literature [@ruscio2004boundaryissues; @ruscio2004checklist].
+Admixture in the reviewed studies ranged from sampling participants through different methods, only some of which explicitly targeted clinical populations [@james2016latent], to combining clinical and non-clinical samples [@deserno2023probing; @frazier2023asd].
 In one case, inclusion criteria from a larger database were partly unclear, but the sample was explicitly structured by diagnostic status [@deserno2023probing].
@@
 non-phonological).
-However, the authors themselves acknowledged this possibility, writing that *"depending on the dyslexia criterion used, samples may vary in composition with regard to their subtypes"* (p.33).
+However, the authors themselves acknowledged this possibility, writing that *“depending on the dyslexia criterion used, samples may vary in composition with regard to their subtypes”* (p.33).
+A subtle and potentially overlooked cause of admixture in studies targeting subtypes within entirely clinical samples is that the original diagnoses might be based on exceeding cutoff thresholds on partly independent dimensions/scales, reflecting a variant of the second type of problematic sampling listed by @ruscio2004checklist In general, a taxonic structure of subtypes should be regarded as inconsistent, and in principle unlikely, if the main condition as a whole is seen as dimen
 
@@
 Only two notable mismatches were identified: @marcus2012latent reached a dimensional conclusion regarding ADHD despite having incurred admixture of clinical cases, while @munson2008asd reached a taxonic conclusion about ASD subtypes without evidence of admixture (all children had received an ASD diagnosis based on a seemingly uniform clinical assessment).
-Together with the above discussed exception in a secondary analysis of a subsample by @frazier2010asd targeting ASD, these findings reinforce the view that ADHD is best conceptualized as dimensional, while ASD may exhibit taxonic features.
+Together with the above discussed exception in a secondary analysis of a subsample by @frazier2010asd targeting ASD, these findings reinforce the view that ADHD is best conceptualized as dimensional, while ASD may exhibit certain taxonic features.
 However, the latter conclusion is supported by limited unbiased evidence.
 
-In our simulated example, we found that artificial admixture represents a major threat to the validity of taxometric analysis, and to clustering methods as well.
+In our simulated example, we showed that artificial admixture represents a major threat to the validity of taxometric analysis, and to clustering methods as well.
 One might argue that our simulated scenario, in which half of sample was separately recruited, was extreme.
 However, this design is broadly consistent with the degree of admixture observed in the reviewed literature.
-It is worth noting that admixture is not limited to the obvious case of oversampling diagnosed individuals, and sometimes occurs in subtle ways.
-For example, by selectively drawing subsamples based on alternative clinical criteria within a larger dataset, as is often the case in subtype studies [see "subtractive" compound sampling, @haslam2020metaanalysis].
 
-An important issue related to the present discussion concerns how we should treat traditional diagnostic categories such as those found in the DSM-5, which have recently come under criticism [@astle2022trandiagnostic; @sonuga2020kuhnian; @sonuga2022dontfit].
-We argue that the question of whether clinically relevant conditions reflect discrete categories or multivariate, correlated continua does not necessarily undermine the practical utility of diagnostic systems.
-While the DSM-5 may seem to endorse a categorical model, it remains atheoretical and does not explicitly or scientifically commit to a taxonic view.
-Researchers, rather than diagnostic manuals, may risk reifying these categories as natural taxa, overinterpreting their conceptual status.
-Hence we advocate for more high-quality taxometric studies.
+## Conclusions, Limitations, Future Directions
 
-## Limitations and Future Directions
+An obvious limitation of the present review concerns the small number of studies identified, and the virtual lack of studies exploring taxa within learning disorders.
+Furthermore, the available evidence was too scattered and methodologically heterogeneous to conduct a meta-analysis.
+These limitations are themselves informative, highlighting the need for more taxometric investigation in the field.
+It is unlikely that many more studies would be included by relaxing eligibility criteria.
+Exclusion criteria specified the lack of use of behavioral indicators; however, no study was excluded on this basis.
+One record was screened out for being a dissertation [@clemons2006taxometric]; however, the full document could not be retrieved.
+Based on its abstract, it focused on whether ADHD subtypes represent distinct taxa, and it reached a taxonic conclusion, but some degree of admixture was possible (as participants were separately recruited from clinical centers and schools).
 
-A first and obvious limitation of the present review concerns the small number of studies identified, and the virtual lack of studies targeting learning disorders as taxa.
-This prevented us from conducting a meta-analysis: the available evidence was too scattered and methodologically heterogeneous.
-However, this limitation is itself informative, underscoring the need for more taxometric investigation in the field.
-More studies might have been included by relaxing the eligibility criteria.
-However, this is unlikely to have made a substantial difference.
-Exclusion criteria specified the lack of use of behavioral indicators, but no study was excluded on this basis.
-One record was screened out for being a dissertation [@clemons2006taxometric].
-The full document could not be retrieved.
-Based on its abstract, it is clear that it focused on whether ADHD subtypes represent distinct taxa, and it reached a taxonic conclusion.
-However, some degree of admixture was possible (participants were separately recruited from clinical centers and schools).
+A relevant future direction concerns a systematic examination of the use and results of cluster analysis, which was employed by some of the reviewed articles, and is a prominent tool in this field [@astle2022trandiagnostic].
+Model-based approaches, such as latent class/profile analysis, were the most frequently used.
+Clustering algorithms are not without problems.
+First, while they are more flexible and can address broader questions, they do not offer direct or consistent tests of taxonicity [@beauchaine2003taxometrics; @ruscio2004boundaryissues].
+Second, meaningful scientific inference from clustering (beyond its use as a mere data reduction technique) requires careful handling of assumptions, as the risk of detecting pseudoclusters is high [@toffalini2022entia; @toffalini2024clusters].
+Finally, it is not immune to risks such as artificial admixture.
+Nonetheless, clustering and taxometric analysis could be more directly, and critically, compared when addressing the question of whether neurodevelopmental conditions reflect categories or dimensions.
 
-A relevant future direction for this line of investigation concerns a systematic examination of the use and results of cluster analysis.
-Indeed, some of the reviewed articles employed clustering methods, and clustering is a prominent tool in this field [@astle2022trandiagnostic].
-Model-based approaches, such as latent class/profile analysis, were the most frequently used.
-The use of clustering is not without problems.
-First, while it is more flexible and can address broader questions, it does not offer a direct or consistent test of taxonicity [@beauchaine2003taxometrics; @ruscio2004boundaryissues].
-Second, meaningful scientific inference from clustering (beyond its simple use as a data reduction technique) requires careful handling of assumptions, as the risk of detecting pseudoclusters is high [@toffalini2022entia; @toffalini2024clusters].
-Finally, it is is not immune to risks such as artificial admixture.
-Nonetheless, clustering and taxometric analysis could be more directly, but also critically, compared when addressing the question of whether neurodevelopmental conditions reflect categories or dimensions.
+A crucial implication of our work concerns how taxometric findings should inform our interpretation of traditional diagnostic categories of NDCs such as those in the DSM-5, which have come under criticism [@astle2022trandiagnostic; @sonuga2020kuhnian; @sonuga2022dontfit].
+Although described as atheoretical and agnostic about etiology, the DSM operationalizes disorders in ways that reinforce a neo-Kraepelinian medical paradigm, implicitly reflecting the idea that NDCs reflect discrete categories underpinned by well-defined, specific dysfunctions.
+This view continues to shape research and clinical practice [@coghill2012categories].
+Finding that the latent structure of one or many NDCs is dimensional does not deny the practical utility of diagnostic systems.
+However, it requires researchers and clinicians to be explicit about their assumptions and models, instead of implicitly reifying categories as natural taxa and overinterpreting their ontological status.
+High-quality taxometric studies offer a direct solution of the issue.
+We argue that they deserve a more central role in the study of NDCs.
 
